---
title: "A systems-modelling approach to predict biological responses to extreme heat"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: \usepackage{bm}\usepackage{lineno}\linenumbers
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

Daniel W.A. Noble^1^ $\ddagger$, Margaret Mayfield^2^ $\dagger$, Ary Hoffmann^1^ $\dagger$, Zhong-Hua Chen^3^ $\dagger$, Steven Lade^5^ $\dagger$, Xuemei Bai^5^ $\dagger$, Danielle Way^4^ $\dagger$, Belinda Medlyn^3^ $\dagger$, Adrienne Nicotra^1,4^ $\dagger$, James Cook^3^, Brajesh Singh^3^, Brendan Choat^3^, Owen Atkin^4^, Barry Pogson^4^,  Alison Bentley^4^, Ian Wright^3^, Micheal R. Kearney^2^ $\ddagger$
 
**Affiliations**:

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra ACT 2602, Australia \   
^2^ School of Biosciences, The University of Melbourne, Australia\ 
^3^ Hawksbury Institute for the Environment, Western Sydney University, Australia\
^4^ Division of Plant Sciences, Research School of Biology, The Australian National University, Canberra ACT 2602, Australia\
^5^ Fenner School of Environment and Society, The Australian National University, Canberra ACT 2602, Australia\ 

^1^ $\ddagger$ Correspondance: daniel.noble@anu.edu.au & m.kearney@unimelb.edu.au\
$\dagger$ These authors contributed equally to this work.

# Abstract (Dan)
Anthropogenic climate change is leading to more frequent and extreme heat waves. These short-term but large-scale events are radically re-shaping interactions among organisms, impacting biodiversity, community composition and ecosystem services crucial to natural systems and food security. Predicting heat wave impacts on interacting species requires an understanding of the processes driving differential exposure and sensitivity of organisms to extreme heat events in a life-cycle context. To achieve this predictive capacity, we need to integrate models across scales while capturing species-specific responses at the level of individuals. We show how existing models in disparate fields can be linked to achieve a level of understanding necessary for calculated responses to extreme heat from individuals to socioecological systems, now and into the future.

# A need for a mechanistically informed modelling framework for interpreting and predicting the effects of extreme heat on community assemblages (Dan)

Climate change is leading to warmer and more variable thermal environments globally [@Barriopedro2023; @Aglas-Leitner2024]. Greater thermal variability will result in organisms experiencing extreme heat waves that result in thermal stress that impacts organismal growth, survival and reproduction with cascading effects on population dynamics, community composition and ecosystem structure and function [@Reichstein2007; @Margalef-Marrase2020]. Climate variability has already been linked to dramatic declines in plant and animal performance, resulting in reduced pollinator abundance [@Kazenel2024] and in crop yields globally [@Zhao2017; @Tito2018; @Bastos2014], but the causes underlying such declines, and their ramifications through communities, are not fully understood. The impacts of extreme heat on species and communities are driven by a combination of direct effects of heat stress on organism physiology and fitness [@Hesketh2023; @Martinez2023; @Rezende2020; @Orsted2024] and indirect effects on interactions (both positive and negative) among species [e.g., @Lopez-Goldar2024; @Huynh2024] -- including the microbial communities many species rely upon [@Trivedi2022]. Understanding the dynamic interplay between direct and indirect interactions, and how these are mediated by environmental factors, has been hampered by the complexity of community interactions and the lack of an integrated modelling framework that can translate short-term physiological damage into changes in fitness and population growth across species [@Trivedi2022].

Predicting how extreme heat impacts population dynamics and community interactions requires the coupling of biophysical, physiological, population and community ecology models. We argue that such a feat is becoming increasingly tangible and will provide powerful new ways of predicting how extreme heat shapes community dynamics. Physiological models can now be seamlessly integrated with biophysical models of heat and mass exchange to characterize temperatures experienced by organisms and simulate the effects of extreme heat events on the entire life cycle of species within communities [@Kooijman2010; @Kearney2009b; @Jusup2024; @Kearney2024; @Briscoe2023]. Physiological models that incorporate estimates of thermal sensitivity across species can capture the delicate balance between damage and repair in physiological systems [@Orsted2024; @Rezende2020; @Rezende2014] to translate the immediate and cumulative impacts of extreme heat on growth, survival and reproduction. Importantly, mechanistic physiological models provide common currencies that can be integrated into population and community ecology models to capture how extreme heat events perturb interactions among species [@Bimler2018; @Bowler2022].

Here, we bring together and discuss modelling approaches across diverse fields to establish a bridge for thinking about how we might begin integrating them to predict how extreme heat will affect communities. We focus our discussion on plant, insect and microbial communities given the strong interconnections between them, the ease with which they can be studied, and their significance to natural and agricultural productivity. Importantly, ‘extreme heat’ can take on a different meaning for these diverse organisms. For example, plants and animals have different thermal tolerances or can exploit microthermal environments in different ways to dampen the effects of extreme atmospheric heat events. As such, ‘extreme’ will be relative to the biology of the organism in question. Our framework can capture these complexities and be extended to incorporate other, diverse interactions. We highlight the immense opportunities such an integration may provide through a greater understanding of the ways in which extreme heat stress manifests across species. Such an integration will help elucidate how and why interactions among species can change, with implications for natural and agricultural systems, food security and ecosystem services. We discuss the challenges of applying such an integrated modelling framework and how it can incorporate other critical features that capture evolutionary change and plasticity.  

# A systems-modelling approach to modelling biotic impacts of heatwaves 

We propose a systems-modelling approach to model biotic impacts of heatwaves (@fig-sys), scaling from the level of the atmosphere to microclimate, and from individuals to populations and communities. Ultimately, the impacts of extreme heat events on a community are driven by responses of individuals. Therefore, to understand extreme heat impacts on a community of organisms, we must characterize the proximal thermal environments – the ‘microclimates’ – experienced by individuals of different species under a given atmospheric event. These microclimates must then be translated into physiological and behavioral responses, including the processes of thermal damage and repair, and how individuals mitigate their exposure to extreme heat through habitat selection and other avoidance strategies. Finally, dynamic energy budgets translate these individual responses into growth and reproduction trajectories across life cycles to drive population and community models of the resulting interactions. In the next sections, we outline new developments that enable us to model all the components in @fig-sys such that a truly systems-based approach to modelling biotic impacts of heatwaves is now possible. 

```{r, fig-sys}
#| label: fig-sys
#| dev: png
#| eval: true
#| fig-cap: "The proposed modelling integration developed here integrates (1) climate/biophysical models to predict the temperatures species experience in their respective microhabitats. (2) Temperatures experienced by organisms provide inputs for damage-repair and dynamic energy budget models that integrate temperature exposure with explicit physiological models that simulate development, behavior, survival and reproduction in response to temperature across life history stages. Physiological models can be built around the unique life cycles of the diverse species in communities, capturing lagged responses to extreme heat, phenological mismatches and mechanistically informed responses to extreme heat for species of the system in question. Outputs (behavior, growth rate/ biomass accumulation, survival and reproduction) from physiological models of individual organisms can then be integrated into (3) population and community ecology models to predict population growth and community composition under a specific type of change. We discuss how (4) environment-mediated feedbacks influence organism thermal exposure and sensitivity (e.g., water availability, microbiota) which can vary across different life stages and organs. The integration of existing models in this novel way will allow simulations at the organismal level to generate emergent predictions for entire communities. In turn, such simulations will allow quantitative predictions to be made on how extreme heatwaves perturb biological systems and allow for the development and implementation of strategies to enhance both biological and socioecological system resilience to extreme heat."

pacman::p_load(magick, ggplot2, patchwork, here, grid)

# Load the image
img <- image_read(here("figs", "Fig1.png"))
grid::grid.raster(img)

```

# The current climate of extreme weather modelling

An understanding of heatwave impacts on biology requires environmental inputs about weather conditions. These data may come from weather stations or models. Weather station observations of air temperature, radiation, wind speed, cloud cover and humidity are readily available as raw timeseries from the stations themselves, and as interpolated products such as WorldClim [@Fick2017]. 

Weather stations can of course only tell us about past heat wave events. Climate modelling and weather event forecasting is achieved through General Circulation Models (GCMs) – large partial differential equation networks of atmospheric, oceanic and biospheric energy and mass exchange processes. Future climate change scenarios are generally produced at monthly time-scales, but can be superimposed on historical weather patterns to infer the nature of future heatwaves (**Belinda – add something more here?**). Historical analyses are facilitated by GCMs via ‘reanalysis’ datasets produced by calibrating GCMs with observations to produce continuous gridded weather data, including ‘surface’ temperatures, soil temperatures and profiles of temperature, pressure, wind speed and humidity at different levels in the atmosphere [@Kalnay2018; @Hersbach2020; @Klinges2022; @Meyer2023; @Munoz-Sabater2021]. The most state-of-the-art of these are ERA5 and ERA5-Land, hourly products at ~31km and ~9km resolutions, respectively. 

Historically, GCMs were coded in fast but inflexible languages like Fortran and C. The emergence of the Julia language [@Bezanson2012] for scientific computing has created an opportunity to rebuild climate models in a more modular manner, allowing easier interfacing between discipline-specific models and easier integration of the massive emerging power of AI for data assimilation and machine learning [@Yatunin2025].  The translation of GCMs and associated land-surface models into more accessible and modular languages opens up new opportunities for collaboration between biologists and earth scientists to better integrate the processes in the atmosphere and the biosphere to generate better predictions of weather events and their biological impacts. 

# From weather to microclimates: predicting community wide exposure to extreme heat 
Extreme weather events are experienced differently across species in a community. For example, while air temperature may be 50°C, an organism a metre below the soil surface may be 25°C cooler (@fig-sys). Exactly how much the soil lags and dampens the atmospheric dynamics depends on the soil properties, especially its moisture level. Conditions above ground depend  on the nature of the vegetation, especially through shading, changes in wind profiles and evaporative cooling (transpiration rates). Vegetation and soil also interact strongly. Past soil moisture regimes and heat wave events may have led to leaf loss and hence increased radiation exposure of the soil surface, while stomatal behavior alters transpiration and hence soil moisture conditions. Strong ‘memory’ effects of past weather conditions can determine how a given heat wave manifests through a habitat, so the same atmospheric heat wave conditions can have different implications for microclimatic conditions depending on the recent history of rainfall and temperature. Despite the complexity and non-linearity of these interactions across the soil-plant-atmosphere continuum (SPAC), there is existing capacity, and growing potential, to model them and determine individual-specific exposure to extreme heat. 

To some extent, the GCMs described in the previous section have these dynamics built into them via modules called ‘land-surface models’ (LSMs), which aim to capture feedbacks between the land surface and associated vegetation. These LSMs will at minimum represent air, soil and canopy temperatures. LSMs that represent the plant canopy with multiple layers will represent the change in leaf temperature with canopy depth; models may also represent thermal gradients through the soil. Land surface schemes will typically do this using energy balance approaches. In contrast dynamic vegetation models may use simpler empirical regression approaches to estimate soil and understory temperatures [e.g. @Braziunas2025].  

The primary role of LSMs is to better understand overall atmospheric dynamics rather than resolve the details of microclimates to which organisms are exposed. Thus, agronomists and ecologists have developed microclimate models that can provide more detailed inferences of environmental exposure. These microclimate models take the atmospheric conditions as independent forcing variables, together with detailed information on terrain (slope, aspect, hill shade), vegetation [leaf/plant-area index, leaf area index (LAI) and soil properties (thermal and hydraulic properties)] to predict how radiation, wind speed, air temperature, humidity, soil temperature and soil moisture vary on centimeter scales. The available models vary in the processes they incorporate and the extent that they use physical first principles [@Kearney2017; @Maclean2019; @Kearney2020; @Maclean2021; @Buckley2023; @De_Frenne2024; @Levy2016]. Our [*Supplementary Online Material*](https://daniel1noble.github.io/thermal_tol_interactions/) provides a tutorial on how to use the *NicheMapR* model [@Kearney2017; @Kearney2020] to predict microclimates from climate data.

The above advanced create opportunities to more closely couple the dynamics of the atmosphere and vegetation with microclimate models. In particular, plant water use dynamics during heatwaves can amplify or reduce high temperatures via effects on photosynthetic capacity, stomatal decoupling, cuticular conductance, leaf damage and plant mortality. These impacts obviously are important for the plants themselves, but also for thermal regimes within the plant canopy. Thus, forecasting future thermal regimes also involves forecasting future vegetation dynamics, with interesting challenges that include interactions between high LAI and canopy shading [@Manzoni2022].

# Translating exposure to organism temperature: biophysics to the rescue!

Once microclimates are quantified, it is now straight-forward to compute heat-budgets of organisms as a function of their traits using the principles of biophysical ecology. Biophysical models have a long history [e.g. @Porter1973; @Porter1969] but have become more widely applied in ecology in the past 20 years, facilitated by developments in environmental datasets, microclimate modelling, and the emergence of high-level programming languages such as R [@Briscoe2023]. Biophysical models of ectothermic animals can account for complex radiative heat transfer and the role of evaporative water loss from the skin, breath and eyes [@Kearney2017; @Riddell2017]. Equivalent models of leaves incorporate the dynamic role of stomata [@Duursma2015; @Kearney2024], while models of endotherms can account for the role of insulation and metabolic heat production [@Briscoe2022]. Future development in biophysical modelling of organismal temperature under extreme heat involves understanding the nuances of plastic physiological and behavioral responses. For example, lizards may pant when exposed to high temperatures [@Loughran2020], birds may ‘wind surf’ or seek thermal micro-refugia [@Sharpe2022], and stomatal behavior may depart from the typical responses to vapor pressure and light intensity to allow emergency cooling [@Drake2018], though this response may depend on prior soil moisture conditions [@Posch2024]. The application of biophysical heat budget models in concert with microclimate models can thus be used to infer differential exposure of members of a community to specific temperatures to predict relative exposure with different degrees of detail, from rank-order stress exposure to hourly trajectories that can be linked to physiological models of damage and repair. 

# Capturing the balance between physiological damage and repair to predict multi-species thermal sensitivity (Dan, Adrienne)

Translating how heat waves impact plants and animals not only depends on modelling temperature *exposure* but also the varying *sensitivity* of organisms to extreme heat [@Rezende2014; @Orsted2022]. Sensitivity to extreme heat can be captured by thermal load sensitivity (TLS) / thermal death time (TDT) models that explicitly account for how heat stress depends on both temperature and the duration of time at a given temperature [@Rezende2014; @Arnold2025]. Typically, these models focus on endpoints that include survival (e.g., lethal temperatures, $LT_{50}$ or $LT_{80}$) or some measure of fertility, but this need not be the case [e.g. Table 1 in @Jagadish2021 and @Arnold2025]. Endpoints are predicted by assuming the effect of time at a given temperature decreases survival or fertility exponentially [@Rezende2014; @Orsted2024; @Orsted2022; @Rezende2020]. For example, Ørsted *et al.* @Orsted2024 show that both mortality and fecundity follow a clear exponential relationship with time, with survival and fecundity being compromised most for long thermal exposures at high temperatures in *Drosophila suzuki*. Importantly, Ørsted *et al.* @Orsted2024 also show that reproductive sensitivity can be higher than mortality with heat injury accumulating faster to impact productivity earlier than survival. TLS theory has also been shown to apply well to photosynthetic function in plants [@Jagadish2021; @Faber2024] highlighting the potential generality of the TLS framework. Thermal load sensitivity models explicitly acknowledge that thermal sensitivity depends on the relative accumulation of damage to cellular and sub-cellular systems that compromise physiological function. Without periods of recovery, where damage can be repaired, organisms accumulate damage over time that reduces growth, survival and reproduction. 

The concept of damage/repair is important for understanding how heat waves affect organisms because both processes can result in lagged responses to heat stress – a common feature of extreme heat events [@Martinez-De_Leon2024; @Arnold2025]. Once the thermal sensitivity (i.e., the slope $z$, Figure 1) for a particular endpoint has been quantified it can be used to predict the accumulation of damage during stressful temperatures experienced under realistic conditions [@Orsted2024; @Orsted2022; @Rezende2020; @Arnold2025]. Several approaches can be used to capture the accumulation of damage [@Orsted2024; @Orsted2022; @Rezende2020], but a simplistic way can be computed following Ørsted *et al.* @Orsted2024:

$$ 
HI = \sum_{i = 1}^{T_{e} > T_{c}} \frac{100 \cdot \left(t_{i+1} - t_{i} \right)}{10^{\left(- \frac{1}{z} \cdot max\left(T_{i}; T_{i+1}\right) + \alpha \right)}} 
$$ {#eq-hi}

@eq-hi summarizes heat injures (HI) sustained across time, $t$, during heat stress events (i.e., when body temperatures are greater than the critical temperatures, $T_{e} > T_{c}$) using information about the maximum temperature, $T$, experienced between time point $t_{i}$ and $t_{i+1}$ and the thermal sensitivity (i.e., $z$ and $\alpha$). Once an organism accumulates damage to reach a given threshold the endpoint has been reached. @Arnold2025 provide a framework for integrating damage and repair processes into these models more explicitly. Modelling the intricate balance between damage and repair for a suite of different species can help identify susceptible species, life stages and tissues that are most at risk from extreme heat events, which allows for more accurate predictions of the varying sensitivity of species across a community.

::: {.callout-note icon="false" appearance="minimal"}

# Box 1: Sensitivity to extreme heat is mediated by environmental feedbacks

Natural and anthropogenic environments shape thermal sensitivity and should be considered in any systems modelling approach. For example, drought, often associated with periods of extreme heat, can reduce thermal tolerance in plants by reducing influencing their ability to cool through transpiration leading to leaf damage []. Extended periods of water stress (e.g., several days or more) have also been shown to increase thermal sensitivity in insects, reducing the time they could deal with extreme heat [@Youngblood2025]. 

Heat tolerance of animals and plants can also both be increased and decreased through changes in the gut microbiome and/or through associated microbes within or around tissues. Interactions between plants and microorganisms, such as plant growth-promoting rhizobacteria (PGPR), arbuscular mycorrhizal fungi (AMF), and bacterial or fungal endophytes are known to impact growth, defense, and heat tolerance in plants [@Chouhan2023; @Liu2025; @Marquez2007]. These interactions are usually achieved by stimulating the production of essential phytohormones, secondary metabolites, organic acids, and amino acids that helps the plant to mitigate heat stress [@Huang2019; @Dastogeer2022]. For example, *Bacillus subtilis* residing in the gut of the nematode *Caenorhabditis elegans* provides protection against heat shock as reflected by effects on fecundity [@Hoang2019]. Moreover, this protective effect can influence the ability of the host to evolve further resistance to the heat shock and harboured more bacteria [@Gehman2019].



:::

# Dynamic Energy Budget (DEB) models that capture the effects of extreme heat across life (Mike and Dan)

Dynamic Energy Budget (DEB) models are physiologically explicit life-cycle models of energy and mass uptake and conversion that predict key life-history transitions, growth, reproduction and survival through time [@Kooijman2010; @Nisbet2000]. DEB models consider organisms as a thermodynamic system (adhering to energy-mass balance) capturing the exchange of food, water, respiratory gases and metabolic waste throughout the life cycle [@Briscoe2023].  DEB models can translate short-term (hourly) variation in factors, such as temperature, toxins (e.g., toxicants, [@Jager2020 @Haberle2025]) and resource availability (nutrients) [@Kearney2013; @Kearney2010] into long-term (days, months, years) effects on development, growth, reproduction and survival [@Briscoe2023; @Kearney2009b], though they have been applied to plants much less than for animal and microbial systems[@Russo2022]. DEB theory, when combined with biophysical models more generally, allows the full water budget to be computed to account for heat stress impacts of a hydric nature, and computes obligate heat production of metabolism to feed back to the calculation of body temperature  [@Briscoe2023]. Perhaps most importantly, DEB theory calculates the life cycle trajectory of an organism and so can account for the effects of timing of heat wave impacts relative to life stage. Applied to multi-species assemblages, DEB models can capture how different species’ life-cycles are impacted by heat and interact with each other to predict how traits such as birth, age at maturity, reproductive events, energy flow and lifespan vary across species. While DEB models are parameter rich, key parameters have been estimated for > 5000 species already [@Marques2018; @Kooijman2021] and approaches for using DEB theory to model plant growth and performance are being developed [@Russo2022]. New analytical approaches using phylogenetic imputation methods allow for models to be developed independently for all species even when data are lacking. These methods show promise in predicting DEB parameters for over 1.3 million species [see, @Bruggeman2009]. In addition, software packages such as *NicheMapR* [@Kearney2017; @Kearney2020a] allow for simulations of life-cycles using parameters for diverse species. From these models key outputs on organisms such as body mass and size, reproductive success and events, along with survival can be used to parameterize vital rates needed for population and community models.

# Mechanistic approaches that 'speak' to population and community ecology under extreme heat (Margie) 

In the plant community and coexistence literature, the maintenance of community-level plant diversity is typically modelled using individual fitness models to parameterise population growth models. The most common Individual-based performance models used for this purpose in plants are the Ricker [@Ricker1954a] and Beverton-Holt [@Beverton2012] models. Both of these models assume that species interactions are competitive in nature but the Ricker model allows interaction outcomes to be positive as well as negative, as long as the net outcome of interactions are negative. This is an important feature as ecological theory tells us that facilitative interactions between species, in which one or both species benefit from the presence of the other, are more common under stressful, or harsh, environments [@Bertness1994].  In thinking about thermal extremes, it is likely that species experience a switch from amenable to harsh conditions, making it likely that interaction outcomes may flip signs [@Bimler2018].  

Based on modern coexistence theory, we can calculate the mutual invasibility of two populations while interacting with each other to determine if each pair of species in a community can persist indefinitely.  This approach typically does not work for predicting whole community diversity, but the tools developed to test for coexistence are well suited to exploring how to link biophysical models with population growth models to advance out ability to predict how extreme thermal events will impact species in the long-term within a community context.  

Recent efforts to increase the biological realism of coexistence models have focused on exploring ways to add environmental variation into these models.  Basic coexistence models are parameterised using average fitness estimates for an average environment.  Approaches for integrating demographic and environmental variation into these models are emerging [@Bimler2018; @Bowler2022].  The calculation of technical coexistence is not necessary for these models to help us understand the impacts extreme heat will have on individual species within a community context.  Understanding how population growth will be impacted by the biophysical and physiological consequences of extreme heat events are conceptually easy to add to Ricker and Beverton-Holt population growth models, with new Bayesian frameworks [@Bowler2022] also allowing us to calculate probabilities of population growth and coexistence of population experiencing these conditions.  The simplest approach to this is to modify the vital rate functions used in these models to compare population growth predictions for species that have and have not been exposed to extreme heat during core life history stages.   Physiological models can be used to estimate changes to fitness resulting from an extreme heat event. We can then model the difference in population growth rates for populations that have and have not experienced those events. We can increase complexity by following Bimler *et al.* [@Bimler2018]'s approach to measuring the effects of environmental gradients on the outcomes of pairwise species interactions and the calculation of species fitness and niche differences.  We could use this same technique in combination with biophysical models to estimate the consequences of localized extreme heat of different local extremes on species interactions with each other and the carry-on effects these interactions have on species fitness within local communities.  Taking this a step forward, new methods are now available for calculation interaction networks based on interaction outcomes calculated using Ricker or Beverton-Holt population growth models. Thus we could take these pairwise models of population growth and create whole community networks to assess how heatwaves will alter the structure of community network structures and stability in addition to pairwise interactions [@Bimler2018].  The limitation of these methods is that they are data hungry, but  even this can be handled by emerging approaches for reducing model complexity without losing model power [@Weiss-Lehman2022].

::: {.callout-note icon="false" appearance="minimal"}

# Box 2: Integrating eco-evolutionary feedbacks within a systems-modelling framework 

Extreme environments both impact trait development and act as a potent selective force triggering evolutionary change in populations [@Chevin2017; @Urban2024; @Hoffmann2011]. Evolutionary responses will depend on the amount of phenotypic and genetic variation exposed to natural selection and the strength of selection on phenotypes. Extreme environments pose unique challenges for predicting evolutionary change because they are, by definition, rare events. Adaptive evolution will therefore depend on the frequency of extreme environments and the underlying genetic correlation between traits expressed across environments [@Chevin2017]. In addition, phenotypic plasticity can result in both maladaptive and adaptive outcomes by either shielding selection on genetic variation or by facilitating persistence and selection making the relative role of plasticity and adaptive evolution to building population resilience complex [@Urban2024; @Hoffmann2022; @Ghalambor2015; @Oostra2018; @Noble2019]. 

Plasticity and evolution can be incorporated into a systems modelling framework by allowing environments to govern 'trait' development (plasticity) and by incorporating selection and genetic (co)variance estimates for 'traits' within these models (evolution). We speak of 'traits' broadly here because these need not be traits typically thought of by evolutionary ecologists (e.g., body mass, morphology). Rather, 'traits' can also include the parameters within models (e.g., DEB parameters) that allow functional traits to emerge [@Sousa2010]. Plasticity can be captured by mapping trait development to the environment. In contrast, evolutionary change takes place in multivariate trait space where  genetic variance and covariance between traits (i.e., constraints) along with varying patterns of selection can be used to predcit trait change across generations. One way to make such predictions is by using the multivariate breeders equation [@Lande1979]:

$$
\Delta \bar{\bm{z}} = \bm{G} \cdot \bm{\beta}
$$ {#eq-breeders}

where $\Delta \bar{\bm{z}}$ is the vector of changes in the mean trait values, $\bm{G}$ is the genetic covariance matrix and $\bm{\beta}$ is the vector of standardised selection gradients that regress each trait on relative fitness (i.e., $\bm{\omega} = \bm{\alpha} + \bm{\beta^{T}}\cdot \bm{z} + \bm{e}$; @Walsh2018). While the multivariate breeders equation can be useful for predicting short term evolutionary responses (i.e., one or a few generations) it is likely limited in predictive power over longer timespans because of changes in heritability and selection through time [@Walsh2018; @Hoffmann2011]. 

Measuring and incorporating both plasticity and evolution into our proposed systems modelling framework is challenging because such models usually focus on one or a few ‘traits’. Studies that have attempted to measure plasticity and evolutionary potential in traits demonstrate how such processes can shape outcomes in important ways [@Youngblood2025; @Baeza_Icaza2025; @Kearney2009a; @Ellner2011; @Pottier2022a; @Leiva2024b]. For example, using isogenic lines of *Drosophia melanogaster*, @Leiva2024b show evidence for genetic variation in both upper thermal limits ($CT_{max}$) and thermal sensitivity (i.e, $z$, slope from TDT curves) using a TDT framework. Simulations also show that selection on one parameter will also induce changes in the other. Additionally, thermal sensitivity in *Drosophia* species is also known to respond plastically to temperature [@Baeza_Icaza2025]. In the alpine plant *Wahlenbergia ceracea* despite ample plasticity in response to warming there is little genetic variation in plasticity, thus adaptive evolution of plasticity in response to extreme heat is unlikely [@Arnold2019a]. Indeed, few studies have attempted to quantify genetic variation in plasticity in response to warming, never mind extreme heat events, and among those that have there is little evidence that such plasticity is adaptive [@Arnold2019a; @Arnold2024]. One challenge in assessing the adaptive evolution of plasticity is that trait values themselves are also under selection in the different environments that elicit the plasticity; thus selection on plasticity reflects both selection on trait means in a given environment and selection on the full range of phenotypes across environments [@Arnold2019a; @Lande1979]. 

:::

::: {.callout-note icon="false" appearance="minimal"}

# Box 3: Interacting species interacting with people: social-ecological systems

Thermal stress to ecosystems may threaten the services those ecosystems provide to people [@Assessment_undated-fv]. For example, loss of output from production ecosystems will have substantial economic consequences for the local area. Degradation of natural ecosystems may affect their regulating services, such as flood mitigation, affecting nearby areas. Impacts on cultural services may be severe for communities that are closely connected to a landscape. 

In turn, people may respond to thermal risks to ecosystems in a variety of ways. For example, farmers or ecosystem managers may introduce new species or alter management practices; partial loss of ecosystems may lead farmers to farm remaining production ecosystems more intensively; and thermal stress may lead people to leave those areas where the stress is occurring, potentially leaving remaining communities more vulnerable to the impacts of thermal stress [@Dandy2019]. Thermal stressors may even influence producers’ psychological beliefs about their level of control over land management, and perceived resources available for management practices [@Price2014]. Thermal risks to ecosystems may also precipitate declines in individual wellbeing, and these consequences may accrue differently across regions and across peoples [@Leviston2018], and compound or accelerate the prior responses. Many of these impacts and responses, however, are poorly understood. 

The impacts of thermal stress on ecosystems can affect not only nearby communities but also human dominant systems such as cities far away from the affected ecosystems. Cities are human dominant complex systems that are intrinsically connected to and relying on other regions to maintain their functionality [@Bai2016; @Grimm2008]. On the other hand, adapting thermal risks in upper stream producers may also be influenced by the perception and behavior of downstream consumers [@Hendricks2022]. The cascading [@Sutanto2020] and telecoupled [@Liu2007] impacts on urban systems can be manifested through multiple ways. For example, extreme heat stress may affect essential ecosystem services such as food production (ref), which can in turn affect price and access to food to people living in urban centers, with financial, economic and health consequences. The impacts on local communities could lead to loss of livelihood, which forms a pressure that push rural people to migrate to cities [@Black2011].  These external pressures will interact with a system that is already constraint by multiple stressors, including direct impacts imposed by extreme heat and other aspects of clime change- to maintaining its normal functionality and the health and wellbeing of its citizens.  If sufficiently widespread, economic and political destabilisation may occur as already observed in other parts of the world [@Neil_Adger1999], which could in turn affect the management of the original ecosystems. 

Understand these relationships between people and nature requires adopting a social-ecological systems perspective, where people and nature are interdependent and intertwined [@Folke2016]. Social-ecological models aim to capture the social systems, natural systems and their interactions [@Schluter2012]. Such models can display non-trivial emergent behavior resulting from complex interactions within the social-ecological systems. For example, responding to decreased productivity from thermal stress by farming more intensively could lead to a maladaptation or ‘lock-in’ from which it is difficult to escape [@Wise2014]. 

Social-ecological models are conventionally built using process-based modelling frameworks such as agent-based models, dynamical systems models or state-and-transition models [@Biggs2021]. Process-based approaches are critical to understanding how emergent social-ecological phenomena develop [@Brown2021], such as tipping points, traps or other dynamics. Constructing process-based dynamical models requires knowledge and data about: the biophysical processes, which could be provided by the models from the previous sections; social processes, which could be developed from a combination of behavioral theory and empirical data; and interactions between them in the system of interest. Since obtaining all these data can be challenging, social-ecological modelling is often performed in an exploratory mode [@Kwakkel2013] to anticipate possible future modes of behavior, such as the direction of change of human behavior and environmental state or the risk of emergent phenomena such as tipping points, rather than trying to predict specific outcomes. 

:::

# 'Hot' solutions for a changing world (Alison, Owen and Belinda, Mike)

Extreme heat will have dramatic and widespread effects on the socioecological systems that humans rely upon (**Box 3**). For example, under current global climate change scenarios, each degree Celsius increase in global mean temperature is estimated to reduce the global yield of wheat by 6.0% and soybean by 3.1% [@Shekhawat2022] exacerbating food shortages globally. A systems modelling approach can provide us with quantitative tools to make informed predictions, plan interventions and evaluate various scenarios for mitigating the impacts of extreme heat. For example, microclimate and biophysical models, in combination with watering interventions, can allow us to make informed predictions about how to mitigate heat stress to crops (See Case Study 4 in the [*Supplementary Online Material*](https://daniel1noble.github.io/thermal_tol_interactions/)). Identifying and/or modifying the thermal suitability of landscapes to benefit plants and animals under extreme heat shows immense promise [@Briscoe2023]. Additionally, as our understanding of the mechanisms associated with heat tolerance improve, genetic tools, environmental manipulations and artifical breeding can be used to enhance thermal tolerance creating more resilience to extreme heat events [e.g., @Marquez2007]. Integrating these solutions into a systems modelling framework will then allow for a calculated understanding of the viability of such approaches in real world situations. 

::: {.callout-note icon="false" appearance="minimal"}
# Outstanding Questions
1. How effective are models using parameters of closely related species at predicting the impacts of extreme heat on species and communities lacking data?
2. How can we capture longer-term evolutionary responses to extreme heat in a systems modelling framework?
3. For modular plant species, can we build and couple dynamic energy budget models to predict how extreme heat will impact tissue and life cycle dynamics?


:::

# Concluding Remarks
We have discussed ways in which a better understanding of the impacts of extreme heat on plant and animal systems can be captured through the coupling of models from diverse research fields – models that translate the effects of atmospheric conditions to the biophysical experiences of organisms and how this then scales up to impact individuals, populations and communities. As part of demonstrating the ways in which these models can be coupled, we have developed a series of case studies to show how such a systems modelling framework can be approached (see [*Supplementary Online Material*](https://daniel1noble.github.io/thermal_tol_interactions/)). While our case studies are necessarily hypothetical in nature they move us one step closer to implementing these models in real-world systems. However, there are still many challenges and gaps (see **Outstanding Questions**). Developing a systems modelling framework will require collaboration between different research fields, including biophysics, physiology, ecology and evolutionary biology as well as socio-ecologists. By working collaboratively and integrativly we can develop a more calculated understanding of the impacts that extreme heat will have on organisms, populations, and communities along with how to mitigate these impacts to preserve diverse systems now and into the future.

# References

<br>

::: {#refs}
:::

<br>

```{r, performancecurve}
#| include: false
#| echo: false
#| message: false
#| warning: false
#| error: false
#| eval: false
## See box 1 in Weiss-Lehman et al. 2022 for a description of the model. This could be applied to non-linear fitness effects based on optimal environment.

#parameters
bmax = 2  # Maximum pop growth rate lamda (think of this as optimal fitness)
 zi  = 5  # Optimal environment (the optimal temperature where fitness is maximised)
 sdi = 3  # Niche breadth (this is the variation in the fitness surface)
  xe = seq(-10,10,0.1) # Environmental gradient (self-explanatory)

lamda <- bmax*exp(-((zi-xe)/(2*sdi))^2)
plot(xe,lamda, type = "l", ylab = "Population growth rate (Lambda)", xlab = "Environmental gradient")

## Another way of doing this from Agelettas book pg 229
#parameters
 rm = 1.5  # Maximum pop growth rate lamda (think of this as optimal fitness)
 z  = 35  # Optimal environment (the optimal temperature where fitness is maximised)
 sd = 8  # Niche breadth (this is the variation in the fitness surface)
 tb = seq(0,80,0.1) # Environmental gradient (self-explanatory)
r <- rm - ((z-tb)^2/(2*sd^2))

plot(tb,r, type = "l", ylab = "Fitness", xlab = "Environmental gradient")


```

```{r, skewnormal}
#| include: false
#| echo: false
#| message: false
#| warning: false
#| error: false
#| eval: false
# Skew normal fitness functions from https://academic.oup.com/evolut/article/73/2/145/6882140
#The parameter θ specifies the location, sigma the width, and alpha the skew of the fitness function. α > 0 gives a right-skewed function (positive skewness), α < 0 a left-skewed function, and α = 0 a symmetric Gaussian fitness function. Importantly, α also changes the position of the maximum value of w (despite location parameter θ being kept constant), so we will write θ0 as the value of θ that provides maximum fitness for a trait value of zero.

# Library
pacman::p_load(latex2exp, ggplot2, dplyr, tidyr, viridis)

# delta; Optimal environment (the optimal temperature where fitness is maximised)
# sigma; Niche breadth (this is the variation in the fitness surface)
# alpha; Skewness of the fitness function

# Environmental Temperature
  set.seed(23)
     z = seq(0,45,0.1) # Temperature gradient
     n = 40            # Number of individuals
ind_sd = 0.5           # Standard deviation among individuals

# Simulation 1. Assume optimum is rather constant among individuals 
#parameters
delta <-  35  
sigma <-  9  + rnorm(n, 0, ind_sd) 
alpha <- -15 + rnorm(n, 0, ind_sd)

w<- list()
for(i in 1:length(alpha)){
w[[i]] <- 2*exp(-(z-delta)^2/(2*sigma[i]^2))*pnorm(alpha[i]*(z-delta)/sigma)
}

w2 <- data.frame(do.call(cbind, w))  %>% pivot_longer(everything(), names_to = "run", values_to = "fitness")  %>% mutate(z = rep(z,each = n))

my_theme <- function(){
  theme(legend.position = "none", 
           axis.title.y = element_text(size = 20), 
           axis.title.x = element_text(size = 20),
           axis.text.x = element_text(size = 18),
            axis.text.y = element_text(size = 18))
}

ggplot(w2, aes(x = z, y = fitness, color = run)) + geom_line() + theme_classic() + labs(y = "Physiological Rate", x = TeX("Temperature ($^\\circ$C)"), color = "Individual") + scale_color_viridis_d() + my_theme()+ geom_vline(xintercept = 28, linetype = "dashed", color = "red") +     geom_vline(xintercept = 18, linetype = "dashed", color = "blue")    

```
