---
title: "A systems-modelling approach to predict biological responses to extreme heat"
date: "`r Sys.Date()`"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "index.html"
    theme: cosmo
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
---

# Introduction

In this supplementary material, we provide extended versions, with code, of our case studies we draw upon within our manuscript to demonstrate how a systems modelling approach can be implemented in practice. Our case studies focus on plant-insect-microbe interactions, and how these interactions are influenced by extreme heat events. We provide the code for the models we use in our case studies, and the code for the simulations we run to predict the impact of heatwaves on these interactions. It is important to realize that extreme heat events are relative to the system in question. What qualifies as 'extreme' will vary across species and communities. Models and code will of course always need to be modified to suite the specific system in question and should be validated against empirical data.

# Load packages

Our case studies all use the statistical software [R](https://www.r-project.org/). We use a number of packages to run our simulations, including [`NicheMapR`](https://mrke.github.io) [@Kearney2009b; @Kearney2013], [`microclima`](https://github.com/ilyamaclean/microclima) [@Maclean2019; @Kearney2020], and [`plantecophys`](https://github.com/RemkoDuursma/plantecophys) [@Duursma2015].  The code below loads the required packages that are needed for this tutorial.

```{r, libraries}
#| label: libraries
#| echo: true
#| warning: false
#| message: false

# You will need to install a few packages first, a couple from GitHub as they are not on CRAN. Note that you will need to remove '#' from the code below to install the packages.
# install.packages("pacman")
# devtools::install_github('mrke/NicheMapR')
# devtools::install_github("ilyamaclean/microclima")

# Load required libraries after the installation
pacman::p_load(devtools, deSolve, ggplot2, dplyr, tidyr, purrr, patchwork, equatags, flextable, NicheMapR, terra, microclima, plantecophys, magick, here, ggrepel, lutz)
source(here("R", "func.R"))

```

:::{.column-margin}
:::{.callout-note}
gFortran can be a problem for NicheMapR so you need to set the paths properly if you are getting errors. Follow details [here](https://stackoverflow.com/questions/69639782/installing-gfortran-on-macbook-with-apple-m1-chip-for-use-in-r). The following code will help you set the paths properly.

```{r}
#| echo: true
#| eval: false
#| results: hide
#| code-fold: true

# But do the following:
# 1. Install gfortran using homebrew in the terminal/bash window
   brew install gcc 
# 2. Take note of the path to gfortran. You'll need this for later
    which gfortran

# 3. In home directory of terminal create a Makevars file
   mkdir ~/.R
   touch ~/.R/Makevars
   open  ~/.R/Makevars

# 4. Add the following to the Makevars file. Note that you need to make sure your version of fortran is correct and replace 124.2.0_1 with your version. You can find version in terminal by typing `gfortran --version`. In addition you need to replace the path to gfortran with the path you found in step 2.
    FC = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
   F77 = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
 FLIBS = -L/opt/homebrew/Cellar/gcc/14.2.0_1/lib/gcc/11

#5. Restart R and try to install the package again via devtools. It should now work.
```
:::
:::

# Developing a systems modelling approach to predict biological responses to extreme heat
### Case Study 1: Simulating plant and insect interactions under heatwaves

Case study one explores how feedback between plants and insects can happen in response to extreme heat events. Here, we are interested in how the behaviour of grasshoppers is influenced by changes in plant transpiration which can impact microhabitat temperatures of grasshoppers (@fig-case1). While this is a simple situation it's a useful starting point to understand how a systems modelling approach can be built up to predict the impact of heatwaves on species interactions. 

```{r, fig-case1}
#| label: fig-case1
#| echo: false
#| fig-cap: "Grasshopper and plant interactions under heatwaves. Grasshoppers feeding on leaves of a bush in Calperum Station, South Australia. Complex leaf structure provides shade and cooler microclimate for grasshoppers. Leaves exposed to hot temperatures close their stomata to conserve water reducing transpiration and photosynthesis. Leaves that are not exposed to direct sunlight can keep their stomata open and continue photosynthesis while also transpiring and providing cooler microhabitats that grasshoppers can exploit through thermoregulatory behaviours. Importantly, stomatal behaviour is mediated by drought conditions. More stomata will be closed during drought to conserve water and this will impact the cooling effect of the plant on the grasshopper."

fig_case1 <- image_read(here("figs", "figS1.png"))
fig_case1
```


#### Step 1: Predicting microclimates relevant to plants and animals: A walk through microclimate modelling {#step1}

To predict the impact of heatwaves on plant-insect interactions, we first need to predict the microclimate. Using biophysical models we can use coarse climate, topological, rainfall, elevation and spatial information to make predictions on what the microclimate a species is likely to experience, which can help us predict its likely body temperature (*exposure*). 

:::{.column-margin}
:::{.callout-important}
It is important to recognize that these are predictions. As with any model, they need field validation [@De_Frenne2024]. Nonetheless, they provide an important tool in our systems modelling framework to help us understand how heatwaves are likely to affect species interactions.
:::
:::

To model the microclimate, we can use [`NicheMapR`](https://mrke.github.io) and the [`microclima`](https://github.com/ilyamaclean/microclima) package to predict the microclimate at a given location. 

There are various climate datasets that can be used to predict the microclimate but for simplicity we'll use weather data from 1889 to yesterday from https://www.longpaddock.qld.gov.au/silo/. 

The code below shows how to predict the microclimate at a location close to Renmark, South Australia (Calperum Station). It uses the weather data to calculate various microclimates relevant at various locations and conditions relevant to an organism. For example, it will compute temperatures at depths below ground (e.g., 0 and 50cm at or below the ground as we do here) and above also above ground (20 cm and 1.2 m air temperature above ground). 

```{r, microclimate_soil}
#| label: microclimate_soil
#| echo: true
#| warning: false
#| message: false
#| error: false
#| eval: false  

# First, set the location where NCEAP climate data is located. This is a global dataset. We need this to calculate the microclimate. Only need to do once.
    get.global.climate(folder=here("NCEP_dan", "NCEAP" ))

# Location we want
  longlat <- c(140.79970255974192, -34.159176277782564) # Near Renmark, South Australia
    micro <- micro_silo(     loc = longlat,      # Location
                          dstart = "01/01/2018", # Start year
                         dfinish = "31/12/2019", # End year
                        minshade = 40,           # Minimum shade, %
                        maxshade = 80,           # Maximum shade, %
                          Usrhyt = 0.2,          # Height above ground, m 
                          Refhyt = 1.2,          # Height of reference air temperature, m
                      microclima = 1,            # Use mircoclima 
                              PC = -1500,        # Water potential for stomatal closure
                              SP = 10,           # Stability parameter for stomatal closure
                           email = "daniel.noble@anu.edu.au") # Replace with your email

# Save the microclimate object for later in case we need it again as it avoids needing to rerun.                           
  save(micro, file = here("output", "microclim", "microclimate.RDa"))

```

We can now plot the below ground temperatures to see the thermal profiles at different depths. Organisms can exploit this variation, especially if they can burrow (@fig-soil_temp). We can see that surface temperatures are much more heterogeneous than those at 50 cm below ground. This is because the surface is more exposed to the sun and can heat up more quickly.

```{r, fig-soil_temp}
#| label: fig-soil_temp
#| fig-cap: "Average soil temperature at a location near Renmark, South Australia. Soil temperatures are simulated from the surface (0 cm) to 50 cm below ground (and in minimum shade, 20%) and shown across time (hours) between 01/01/2018 to 31/12/2019. "
#| code-fold: true

# Load the microclimate object
  load(here("output", "microclim", "microclimate.RDa"))

# Explore microclimate data. For ease of plotting we'll extract soil temperatures at the 20% share levels across the full set of dates we requested. 
        soil_temp <- data.frame(micro$soil) # Note this is under 20% shade
  soil_temp$dates <- micro$dates            # Add in the dates

# Re-arrange for plotting
  soil_temp2 <- soil_temp %>% 
                select(D0cm,  D50cm, dates) %>%
                pivot_longer(cols = c(D0cm,  D50cm), names_to = "Depth", values_to = "Temperature")

# Plot soil temps
  ggplot(soil_temp2) + 
              geom_line(aes(y = Temperature, x = dates, col = Depth), show.legend = TRUE) +
              labs(y = "Soil Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
```

If we want to use air temperatures because our organism does not burrow, as in the case for our grasshopper case study (@fig-case1), then we can extract air temperatures at different heights above ground that we estimated (20 cm and 1.2 m) (@fig-Air_temp). Of course, we could calculate temperatures for other heights as well, but this is the space most relevant to our system (@fig-case1). This information is also stored in our 'microclimate' object.

We can simulate microclimates under different scenarios in new calls to the `micro_silo` function but to some extent this can be done within a single simulation. In our case, we've run a single simulation for two heights already so we have everything we need. To obtain our air temperatures, these are stored in a different place in our "microclimate" object. Again, we have access to these calculations under our two shade levels (20% and 80%), but we'll just use our lowest shade level for now. 

:::{.column-margin}
:::{.callout-important}
It's important to recognize that the microclimate object has a lot of relevant information that we can use in our simulations. Keep this in mind while exploring the documentation. We'll come back to this later on.
:::
:::

```{r, microclimate_air}
#| label: microclimate_air
#| echo: true
#| warning: false
#| message: false

# Get the above ground temperatures at 20 cm and 1.2 m
  air_temp <- data.frame(micro$metout)
  air_temp$dates <- micro$dates
```

Now we can have a look at what the air temperature would be 20 cm and 1.2 m above ground, a space relevant to a grasshopper (@fig-Air_temp). Of course, we could simulate temperatures for various heights to factor in movement of the grasshopper in response to heatwaves, but we'll keep it simple for now.

```{r, fig-Air_temp}
#| label: fig-Air_temp
#| fig-cap: "Average air temperature at a location near Renmark, South Australia. Air temperatures at 20 cm and 1.2 m above ground (and in minimum shade, 20%) are shown for a short times between 2019-01-01 and 2019-01-10."
#| code-fold: true

  air_temp2 <- air_temp                             %>% 
                select(dates, TALOC, TAREF)         %>%
                pivot_longer(cols = c(TALOC,  TAREF), 
                         names_to = "Height", 
                        values_to = "Temperature")  %>% 
                mutate(Height = ifelse(Height == "TALOC", "20 cm", "1.2 m"))

# Plot soil temps
  air_temp2  %>% filter(dates > "2019-01-01" & dates < "2019-01-10") %>%
  ggplot(aes(y = Temperature, x = dates, col = Height)) + 
              geom_line() +
              labs(y = "Air Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
              
```

We can immediately see from @fig-Air_temp that air temperatures closer to the ground are hotter than those higher up likely because of increased radiative heat transfer from the ground to air in close proximity and convective cooling higher in the air column.

Interestingly, during these dates we had a major heatwave event that occurred between 2018-12-24 and 2019-01-07. Lets have a look at the temperatures during this period.

```{r, fig-Air_temp_heatwave}
#| label: fig-Air_temp_heatwave
#| fig-cap: "Average air temperature at a location near Renmark, South Australia. Air temperatures at 20 cm and 1.2 m above ground (and in minimum shade, 20%) are shown for a short times between 2018-12-24 and 2019-01-07."
#| code-fold: true

  air_temp2  %>% filter(dates > "2018-12-24" & dates < "2019-01-07") %>%
  ggplot(aes(y = Temperature, x = dates, col = Height)) + 
              geom_line() +
              geom_hline(yintercept = 40, colour = "red", lty = 2) +
              labs(y = "Air Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
```


#### Step 2: Using microclimates to predict body temperatures of plants and animals: example of how cooling effects of plant transpiration impact grasshopper body temperature {#step2}

Taking a systems approach means that we can account for the effects of other organisms on our focal species. To do this, we need to consider the biology of the species that co-exist and form part of the same system. In this case, the plants tha the grasshopper consumes which will influence the microclimate the grasshopper experiences. Plants will also respond to the environmental temperatures and this will directly and indirectly influence the grasshopper.

If this is an grasshopper on a bush as in @fig-case1, the plant itself will also change its transpiration through the day. We can also obtain plant transpiration data from the microclimate model. This can be used to predict the cooling effect of the plant on the grasshopper.

The microclimate model that we ran in [Step 1](#step1) can now be used to predict how the grasshopper and plant will respond to the heatwave. We can then couple these responses to predict the impact of the heatwave on the grasshopper considering how the plant is also responding to heat and impacting teh body temperature of the grasshopper. Before we get temperatures for the grasshopper, we need to get the transpiration data for the plant. 

We can then use this microclimate data to simulate the growth of plants and insects under different temperature conditions [@fig-case1].

Now, lets run the ectotherm model for a grasshopper. This allows us to take in the microclimate and use it to calculate the body temperature of the grasshopper. The ectotherm model also allows us to incorperate relevant behavuioural plasticity in response to extreme heat, which we'll do. 


```{r, ecto_grasshopper}
#| label: ecto_grasshopper
#| echo: true
#| warning: false
#| message: false

# Load the microclimate object. Note ectotherm needs this to run.
  load(here("output", "microclim", "microclimate.RDa"))

# Parameters for the grasshopper
    Ww_g <- 0.2  # g, body mass
   shape <- 1    # cylinder
 shape_b <- 10   # ratio of long to short axis of cylinder
  T_pref <- 38   # preferred body temperature (animal will attempt to regulate as close to this value as possible)
 T_F_min <- 13   # degrees C, lower body temperature for foraging
 T_F_max <- 40   # degrees C, upper body temperature for foraging
 T_B_min <- 12.5 # degrees C, minimum basking temperature
T_RB_min <- 12.5 # degrees C, temperature at which animal will move to a basking site
  CT_max <- 47   # degrees C, critical thermal maximum (animal will die if CT_kill = 1 and this threshold is exceeded)
  CT_min <- 0    # degrees C, critical thermal minimum (used by program to determine depth selected when inactive and burrowing)


# Behavioural traits. This tells the model what the grasshopper can do in response to heat. 
     diurn <- 1     # diurnal activity allowed (1) or not (0)? 
   nocturn <- 0     # nocturnal activity allowed (1) or not (0)?
    crepus <- 1     # crepuscular activity allowed (1) or not (0)?
shade_seek <- 1     # shade seeking allowed (1) or not (0)?
    burrow <- 0     # shelter in burrow allowed (1) or not (0)?
  pct_cond <- 0     # Percentage of animal surface contacting the substrate

# Now that we have set the parameters we can run the model
grasshopper <- ectotherm(Ww_g = Ww_g,
                        shape = shape,
                      shape_b = shape_b,
                       T_pref = T_pref,
                      T_F_min = T_F_min,
                      T_F_max = T_F_max,
                      T_B_min = T_B_min,
                     T_RB_min = T_RB_min,
                       CT_min = CT_min,
                       CT_max = CT_max,
                        diurn = diurn,
                      nocturn = nocturn,
                       crepus = crepus,
                   shade_seek = shade_seek,
                       burrow = burrow,
                     pct_cond = pct_cond
)
      grasshopper_environ <- as.data.frame(grasshopper$environ)
grasshopper_environ$dates <- micro$dates
```

Now that we have run the ectotherm model we can plot the body temperature of the grasshopper over time along side the air temperature. This will allow us to see how the grasshopper responds to the heatwave.

```{r, fig-grasshopper_temp}
#| label: fig-grasshopper_temp
#| fig-cap: "Body temperature of a grasshopper at a location near Renmark, South Australia. Body temperature of the grasshopper is shown in relation to air temperature at 20 cm and 1.2 m above ground (and in minimum shade, 20%) for a short times between 2018-12-24 and 2019-01-07."

grasshopper_environ2 <- grasshopper_environ                                                           %>% 
                        filter(dates > "2018-12-24" & dates < "2019-01-07")                           %>% 
                        select(dates, TA, TC, ACT)                                                    %>%
                        pivot_longer(     cols = c(TA, TC), 
                                      names_to = "Temps", 
                                     values_to = "Temperature")                                       %>% 
                        mutate(Temps = ifelse(Temps == "TA", "Air Temperature", "Body Temperature"))

  ggplot(grasshopper_environ2) + 
              geom_line(aes(y = Temperature, x = dates, col = Temps), show.legend = TRUE) + 
              geom_point(aes(y = ACT, x = dates, fill = factor(ACT)), shape = 21, size = 4) +
              labs(y = "Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
              scale_colour_viridis_d() +
              scale_fill_manual(values = c("0" = "blue", 
                                           "1" = "green", 
                                           "2" = "orange"),
                                labels = c("Inactive", "Basking", "Foraging")) +
              labs(fill = "Behaviour")
              


```

# Using a systems modelling approach ti develop solutions to extreme heat events