---
title: "A systems-modelling approach to predict biological responses to extreme heat"
date: "`r Sys.Date()`"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "index.html"
    theme: cosmo
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
---

# **Introduction**

In this supplementary material, we provide extended versions, with code, of our case studies we draw upon within our manuscript to demonstrate how a systems modelling approach can be implemented in practice. Our case studies focus on plant-insect-microbe interactions, and how these interactions are influenced by extreme heat events. We provide the code for the models we use in our case studies, and the code for the simulations we run to predict the impact of heatwaves on these interactions. Models and code will of course always need to be modified to suite the specific system in question and should be validated against empirical data. We view this tutorial as a starting point to help you understand how to implement a systems modelling approach to predict the impact of extreme heat on species interactions.

# **Load packages**

Our case studies all use the statistical software [R](https://www.r-project.org/). We use a number of packages to run our simulations, including [`NicheMapR`](https://mrke.github.io) [@Kearney2009b; @Kearney2013], [`microclima`](https://github.com/ilyamaclean/microclima) [@Maclean2019; @Kearney2020], and [`plantecophys`](https://github.com/RemkoDuursma/plantecophys) [@Duursma2015].  The code below loads the required packages that are needed for this tutorial.

```{r, libraries}
#| label: libraries
#| echo: true
#| warning: false
#| message: false

# You will need to install a few packages first, a couple from GitHub as they are not on CRAN. Note that you will need to remove '#' from the code below to install the packages.
# install.packages("pacman")
# devtools::install_github('mrke/NicheMapR')
# devtools::install_github("ilyamaclean/microclima")

# Load required libraries after the installation
pacman::p_load(devtools, deSolve, ggplot2, dplyr, tidyr, purrr, patchwork, equatags, flextable, NicheMapR, terra, microclima, plantecophys, magick, here, ggrepel, lutz)
source(here("R", "func.R"))

```

:::{.column-margin}
:::{.callout-note}
gFortran can be a problem for NicheMapR so you need to set the paths properly if you are getting errors. Follow details [here](https://stackoverflow.com/questions/69639782/installing-gfortran-on-macbook-with-apple-m1-chip-for-use-in-r). The following code will help you set the paths properly.

```{r, nichmapr}
#| label: nichmapr
#| echo: true
#| eval: false
#| results: hide
#| code-fold: true

# But do the following:
# 1. Install gfortran using homebrew in the terminal/bash window
   brew install gcc 
# 2. Take note of the path to gfortran. You'll need this for later
    which gfortran

# 3. In home directory of terminal create a Makevars file
   mkdir ~/.R
   touch ~/.R/Makevars
   open  ~/.R/Makevars

# 4. Add the following to the Makevars file. Note that you need to make sure your version of fortran is correct and replace 124.2.0_1 with your version. You can find version in terminal by typing `gfortran --version`. In addition you need to replace the path to gfortran with the path you found in step 2.
    FC = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
   F77 = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
 FLIBS = -L/opt/homebrew/Cellar/gcc/14.2.0_1/lib/gcc/11

#5. Restart R and try to install the package again via devtools. It should now work.
```
:::
:::

# **Predicting relevant microclimates to organisms: A walk through microclimate modelling** {#section1}

Broad-scale climate data is extremely coarse relative to the types of conditions experienced by organisms in their natural environment. Using biophysical models we can use coarse climate, topological, rainfall, elevation and spatial information to make predictions on what the microclimate a species is likely to experience, which can help us predict its likely body temperature (*exposure*). In combination with an understanding of the species' biology we can simulate realistic scenarios that are relevant to the species in question.

As a first step, it's important to be able to down-scale climatic data to microclimate conditions relevant to organisms. To model the microclimate, we can use the [`NicheMapR`](https://mrke.github.io) and the [`microclima`](https://github.com/ilyamaclean/microclima) packages. This section will demonstrate some ways that we can achieve this and make predictions for locations relevant to organisms of interest. 

There are various climate datasets that can be used to predict the microclimate but for simplicity we'll use weather data from 1889 to yesterday from https://www.longpaddock.qld.gov.au/silo/. 

The code below shows how to predict the microclimate at a location close to Renmark, South Australia (Calperum Station). It uses the weather data to calculate various microclimates relevant at various locations and conditions relevant to an organism. For example, it will compute temperatures at depths below ground (e.g., 0 and 50cm at or below the ground as we do here) and above also above ground (20 cm and 1.2 m air temperature above ground). 

```{r, microclimate_soil}
#| label: microclimate_soil
#| echo: true
#| warning: false
#| message: false
#| error: false
#| eval: false  

# First, set the location where NCEAP climate data is located. This is a global dataset. We need this to calculate the microclimate. Only need to do once.
    get.global.climate(folder=here("NCEP_dan", "NCEAP" ))

# Location we want
  longlat <- c(140.79970255974192, -34.159176277782564) # Near Renmark, South Australia
    micro <- micro_silo(     loc = longlat,      # Location
                          dstart = "01/01/2018", # Start year
                         dfinish = "31/12/2019", # End year
                        minshade = 40,           # Minimum shade, %
                        maxshade = 80,           # Maximum shade, %
                          Usrhyt = 0.2,          # Height above ground, m 
                          Refhyt = 1.2,          # Height of reference air temperature, m
                      microclima = 1,            # Use mircoclima 
                              PC = -1500,        # Water potential for stomatal closure
                              SP = 10,           # Stability parameter for stomatal closure
                           email = "daniel.noble@anu.edu.au") # Replace with your email

# Save the microclimate object for later in case we need it again as it avoids needing to rerun.                           
  save(micro, file = here("output", "microclim", "microclimate.RDa"))

```

We can now plot the below ground temperatures to see the thermal profiles at different depths. Organisms can exploit this variation, especially if they can burrow (@fig-soil_temp). We can see that surface temperatures are much more heterogeneous than those at 50 cm below ground. This is because the surface is more exposed to solar radiation and can heat up more quickly.

```{r, fig-soil_temp}
#| label: fig-soil_temp
#| fig-cap: "Average soil temperature at a location near Renmark, South Australia. Soil temperatures are simulated from the surface (0 cm) to 50 cm below ground (and in minimum shade, 20%) and shown across time (hours) between 01/01/2018 to 31/12/2019. "
#| code-fold: true

# Load the microclimate object
  load(here("output", "microclim", "microclimate.RDa"))

# Explore microclimate data. For ease of plotting we'll extract soil temperatures at the 20% share levels across the full set of dates we requested. 
        soil_temp <- data.frame(micro$soil) # Note this is under 20% shade
  soil_temp$dates <- micro$dates            # Add in the dates

# Re-arrange for plotting
       soil_temp2 <- soil_temp %>% 
                     select(D0cm,  D50cm, dates) %>%
                     pivot_longer(cols = c(D0cm,  D50cm), 
                              names_to = "Depth", 
                             values_to = "Temperature")

# Plot soil temps
  ggplot(soil_temp2) + 
              geom_line(aes(y = Temperature, x = dates, col = Depth), show.legend = TRUE) +
              labs(y = "Soil Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
```

If we want to use air temperatures because our organism does not burrow, as in the case for our grasshopper case study that we will describe below (@fig-case1), then we can extract air temperatures at different heights above ground (For example, at 20 cm and 1.2 m) (@fig-Air_temp). Of course, we could calculate temperatures for other heights as well, but this is the space most relevant to our system (@fig-case1). This information is also stored in our `microclimate` object.

We can simulate microclimates under different scenarios in new calls to the `micro_silo` function but to some extent this can be done within a single simulation. In our case, we've run a single simulation for two heights already so we have everything we need. To obtain our air temperatures, these are stored in a different place in our `microclimate` object. Again, we have access to these calculations under two shade levels (20% and 80%), but we'll just use our lowest shade level for now. 

:::{.column-margin}
:::{.callout-important}
The microclimate object has a lot of relevant information that we can use in our simulations. Keep this in mind while exploring the [documentation](https://rdrr.io/github/mrke/NicheMapR/man/). We'll come back to this later on.
:::
:::

```{r, microclimate_air}
#| label: microclimate_air
#| echo: true
#| warning: false
#| message: false

# Get the above ground temperatures at 20 cm and 1.2 m
  air_temp <- data.frame(micro$metout)
  air_temp$dates <- micro$dates
```

Now we can have a look at what the air temperature would be 20 cm and 1.2 m above ground, a space relevant to a grasshopper (@fig-Air_temp). We could simulate temperatures for various heights to factor in movement of the grasshopper in response to heatwaves, but we'll keep it simple for now.

```{r, fig-Air_temp}
#| label: fig-Air_temp
#| fig-cap: "Average air temperature at a location near Renmark, South Australia. Air temperatures at 20 cm and 1.2 m above ground (and in minimum shade, 20%) are shown for a short times between 2019-01-15 and 2019-01-31."
#| code-fold: true

  air_temp2 <- air_temp                             %>% 
                select(dates, TALOC, TAREF)         %>%
                pivot_longer(cols = c(TALOC,  TAREF), 
                         names_to = "Height", 
                        values_to = "Temperature")  %>% 
                mutate(Height = ifelse(Height == "TALOC", "20 cm", "1.2 m"))

# Plot soil temps
  air_temp2  %>% filter(dates > "2019-01-15" & dates < "2019-01-31") %>%
  ggplot(aes(y = Temperature, x = dates, col = Height)) + 
              geom_line() +
              labs(y = "Air Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
              
```

We can immediately see from @fig-Air_temp that air temperatures closer to the ground are hotter than those higher up likely because of increased radiative heat transfer from the ground to air in close proximity and convective cooling higher in the air column.

Interestingly, during these dates we had a major heatwave event that occurred between 2018-12-24 and 2019-01-07. Lets have a look at the temperatures during this period.

```{r, fig-Air_temp_heatwave}
#| label: fig-Air_temp_heatwave
#| fig-cap: "Average air temperature at a location near Renmark, South Australia. Air temperatures at 20 cm and 1.2 m above ground (and in minimum shade, 20%) are shown for a short times between 2019-01-15 and 2019-01-31."
#| code-fold: true

  air_temp2  %>% filter(dates > "2019-01-15" & dates < "2019-01-31") %>%
  ggplot(aes(y = Temperature, x = dates, col = Height)) + 
              geom_line() +
              geom_hline(yintercept = 40, colour = "red", lty = 2) +
              labs(y = "Air Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
                        scale_colour_viridis_d()
```

:::{.column-margin}
:::{.callout-important}
It is important to recognize that these are predictions. As with any model, they need field validation [@De_Frenne2024]. Nonetheless, they provide an important tool in our systems modelling framework to help us understand how heatwaves are likely to affect species interactions.
:::
::: 

# **Developing a systems modelling approach to predict biological responses to extreme heat**
## **Case Study 1: Simulating plant and insect interactions under heatwaves**

Case study one explores how feedback between plants and insects can happen in response to extreme heat events. Here, we are interested in how the behaviour and body temperature of grasshoppers is influenced by changes in plant transpiration which can impact microhabitat temperatures of grasshoppers (@fig-case1). While this is a simple situation it's a useful starting point to understand how a systems modelling approach can be built up to predict the impact of heatwaves on species interactions. 

```{r, fig-case1}
#| label: fig-case1
#| echo: false
#| fig-cap: "Grasshopper and plant interactions under heatwaves. Grasshoppers feeding on leaves of a bush in Calperum Station, South Australia. Complex leaf structure provides shade and cooler microclimate for grasshoppers. Leaves exposed to hot temperatures close their stomata to conserve water reducing transpiration and photosynthesis. Leaves that are not exposed to direct sunlight can keep their stomata open and continue photosynthesis while also transpiring and providing cooler microhabitats that grasshoppers can exploit through thermoregulatory behaviours. Importantly, stomatal behaviour is mediated by drought conditions. More stomata will be closed during drought to conserve water and this will impact the cooling effect of the plant on the grasshopper."

fig_case1 <- image_read(here("figs", "figS1.png"))
fig_case1
```


### **Step 1: Predicting temperatures experienced by plants that form the grasshopper microclimate** {#step1}

To predict the impact of heatwaves on plant-insect interactions, we first need to predict how the microclimate that each organism experiences translates into body temperatures.  Taking a systems approach means that we can account for the effects of other organisms on our focal species. To do this, we need to consider the biology of the species that co-exist. 

In this case, the plants that the grasshopper consumes will influence the microclimate the grasshopper experiences (@fig-case1). Plants will also respond to the thermal environmental, and this will directly and indirectly influence the grasshopper. For example, plants will change their transpiration through the day and this can impact local temperatures experienced by grasshoppers (@fig-case1). 

The microclimate model that we ran in [Section 1](#section1) can now be used to predict how the plant and grasshopper will respond to the heatwave. **Remember, we just used weather data to calculate air temperatures expected at 0.2 and 1.2m in both 20% and 80% shade. We can use this microclimate data to estimate leaf temperatures for a plant and then use the information on plant temperatures to feedback into the calculations of grasshopper body temperature.**

Before we can estimate the temperatures for the grasshopper, we need to use a biophysical model to predict what the plant is doing. We can then use this 'new' microclimate data to simulate the growth of insects under different temperature conditions, changing what would happen if plants respond differently [@fig-case1].

Let's use our microclimate simulations to predict the plant leaf and air temperatures at the relevant heights to grasshoppers.

First, the traits of the leaf must be defined. An ellipsoid shape is used in this simulation. Specification of the mass and the ratios of the linear dimensions of the shape (here, semi-minor axes relative to the semi-major axis) together with the density (default value 1 g cm${^3}$) allows the geometric functions in the ectotherm model to compute the absolute dimensions and surface areas. 

Evaporative exchange is driven by the stomatal conductance parameters, which can be specified for the abaxial and adaxial sides of the leaf as well as the overall base leaf conductance through the cuticle. The `pct_wet` parameter is set to 0.

Radiative exchange parameters can be set including the emissivities of the leaf and the substrate, the solar absorptivities of the leaf, the configuration factors to the sky and ground, the percentage of the leaf conducting to the ground and the orientation to the sun.

```{r, microclimate_plant}
#| label: microclimate_plant
#| echo: true
#| warning: false
#| message: false

  # Load the microclimate object
    load(here("output", "microclim", "microclimate.RDa"))

  # First, lets extract the relevant information from the microclimate object
      PDIFs <- micro$diffuse_frac             # use variable diffuse fraction
        P_a <- get_pressure(micro$elev)       # air pressure, Pa
        P_a <- rep(P_a, nrow(micro$metout))   # create hourly vector
      dates <- micro$dates                    # dates

  # Set leaf functional traits
         live <- 0      # don't simulate behaviour or respiration
         leaf <- 1      # use the stomatal conductance formulation for evaporation
         Ww_g <- 1      # wet weight, g
        shape <- 2      # 0=plate, 1=cylinder, 2=ellipsoid
      shape_b <- 0.0025 # ratio of b axis:a axis for ellipsoid 
      shape_c <- 0.1176 # ratio of c axis:a axis for ellipsoid 
      g_vs_ab <- 0.4    # leaf vapour conductance, abaxial (bottom of leaf), mol/m2/s
      g_vs_ad <- 0.0    # leaf vapour conductance, adaxial (top of leaf), mol/m2/s
  g_vs_ab_max <- 0.4    # leaf vapour conductance, abaxial (bottom of leaf), mol/m2/s
  g_vs_ad_max <- 0.0    # leaf vapour conductance, adaxial (top of leaf), mol/m2/s
  epsilon_sub <- 0.95   # emissivity of the substrate (-)
  epsilon_sky <- 1      # emissivity of the sky (-), accounted for already in 'TSKY' output from microclimate model so make it 1
    epsilon_L <- 0.97   # emissivity of the leaf (-)
      alpha_L <- 0.5    # solar absorptivity of the leaf (-)
       fatosk <- 0.5    # radiation configuration factor to sky (-)
       fatosb <- 0.5    # radiation configuration factor to substrate (-)
 conv_enhance <- 1.4    # convective enhancement factor (-)
     pct_cond <- 0      # percent of leaf conducting to the ground (%)
       postur <- 3      # leaf oriented horizontally, not tracking sun
          M_1 <- 0      # make metabolic rate zero

# Set up vapour conductance vectors and simulate stomatal closure at night
                              f_open <- rep(1, nrow(micro$metout))
  f_open[micro$metout[,"ZEN"] == 90] <- 0      # close stomata when the sun is set
                            g_vs_abs <- g_vs_ab * f_open 
                            g_vs_ads <- g_vs_ad * f_open

# Run the heat-budget model for the plant leaf. Note that the microclimate object is used to get the relevant environmental temperatures to make the calculations of leaf temperature
    plant <- ectotherm(  leaf = leaf, 
                         live = live,
                         Ww_g = Ww_g, 
                        shape = shape, 
                      shape_b = shape_b, 
                      shape_c = shape_c, 
                      g_vs_ab = g_vs_abs, 
                      g_vs_ad = g_vs_ads, 
                  g_vs_ab_max = g_vs_ab_max, 
                  g_vs_ad_max = g_vs_ad_max, 
                  epsilon_sub = epsilon_sub, 
                  epsilon_sky = epsilon_sky,
                      epsilon = epsilon_L, 
                    alpha_max = alpha_L, 
                    alpha_min = alpha_L,
                          M_1 = M_1,
                       fatosk = fatosk, 
                       fatosb = fatosb, 
                 conv_enhance = conv_enhance, 
                     pct_cond = pct_cond, 
                       postur = postur,
                       preshr = P_a, 
                         PDIF = PDIFs)

# Save model
  save(plant, file = here("output", "microclim", "plant_micro.RDa"))

```


```{r, fig-plant_temp}
#| label: fig-plant_temp
#| fig-cap: "Leaf temperature of a plant at a location near Renmark, South Australia. Leaf temperature is shown in relation to air temperature at 1.2 m above ground (and in minimum shade, 40%) between 2019-01-15 and 2019-01-31."
#| code-fold: true
  
  # Load the plant object
    load(here("output", "microclim", "plant_micro.RDa"))
  
  # Extract the microclimate temperatures. Becasue this was simulated for a plot the 'body tempertaure'is the leaf temperature and is stored as 'TC' in the environ object
            environ_plant <- as.data.frame(plant$environ)
      environ_plant$dates <- micro$dates  # For clarirty add in dates from microclimate that ecto model uses

  environ_plant2 <- environ_plant %>% filter(dates > "2019-01-15" & dates < "2019-01-31") %>%
              select(dates, TA, TC) %>%
              pivot_longer(cols = c(TA, TC), 
                           names_to = "Temps", 
                           values_to = "Temperature") %>% 
              mutate(Temps = ifelse(Temps == "TA", "Air Temperature", "Leaf Temperature"))

  ggplot(environ_plant2) + 
              geom_line(aes(y = Temperature, x = dates, col = Temps), show.legend = TRUE) + 
              labs(y = "Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
              scale_colour_viridis_d()
```

We will now use the `plantecophys` package to calculate the leaf temperature and transpiration of the plant. This will allow us to understand how the plant is responding to the heatwave and how this might influence the grasshopper. The `plantecophys` package allows us to get more realistic stomatal behaviour via the `Photosyn` function. This code runs `Photosyn` assuming the leaf temperature calculated with the ectotherm model to start with.

```{r}
#| label: plant_stomatal
#| echo: true
#| warning: false
#| message: false

# Include simulated stomatal conductance from plantecophys
  UMOLPERJ <- 4.57                                                   # mu_mol photons / J
WETAIR_out <- WETAIR(db = plant$metout[,"TALOC"], rh = plant$metout[,"RHLOC"], bp = P_a)
    VPD_Pa <- WETAIR_out$esat - WETAIR_out$e
 photo_out <- Photosyn(Tleaf = plant$environ[,"TC"],                 # Leaf temperature (C)
                          g1 = 5.2,                                  # Parameter of Ball-Berry model
                          Ca = 400,                                  # Atmospheric CO2 concentration (ppm)
                         VPD = VPD_Pa/1000,                          # Vapour pressure deficit (kPa)
                      vpdmin = 0.5,                                  # lower threshold on VPD
                        PPFD = plant$metout[,"SOLR"] * UMOLPERJ * 1, # mu mol m-2 s-1
                        Patm = P_a/1000                              # Atmospheric pressure (kPa)
                      )
g_vs_abs_pho <- photo_out$GS
```

Now recalculate leaf temperature using these stomatal conductance values from `Photosyn`. Note that we have re-written the `g_vs_abs_pho` to capture the stomatal conductance from `Photosyn`. We just replace this in the plant biophysical model.

```{r, plant_stomatal2}
#| label: plant_stomatal2
#| echo: true
#| warning: false
#| message: false

# Rerun the models. Note that the paramaters as the same as  
plant2 <- ectotherm( leaf = leaf, 
                     live = live,
                     Ww_g = Ww_g, 
                    shape = shape, 
                  shape_b = shape_b, 
                  shape_c = shape_c, 
                  g_vs_ab = g_vs_abs_pho, 
                  g_vs_ad = g_vs_ads, 
              g_vs_ab_max = g_vs_ab_max, 
              g_vs_ad_max = g_vs_ad_max, 
              epsilon_sub = epsilon_sub, 
              epsilon_sky = epsilon_sky,
                  epsilon = epsilon_L, 
                alpha_max = alpha_L, 
                alpha_min = alpha_L,
                      M_1 = M_1,
                   fatosk = fatosk, 
                   fatosb = fatosb, 
             conv_enhance = conv_enhance, 
                 pct_cond = pct_cond, 
                   postur = postur,
                   preshr = P_a, 
                    PDIF = PDIFs)
# Save model
  save(plant2, file = here("output", "microclim", "plant2_micro.RDa"))
#environ <- as.data.frame(plant2$environ)
#T_leaf_plantecophys <- environ$TC
```

Now, lets have a look at the leat temperature when stomaltal conduction behaviour is incorperated.

```{r, fig-plant_temp2}
#| label: fig-plant_temp2
#| fig-cap: "Leaf temperature of a plant at a location near Renmark, South Australia. Leaf temperature is shown in relation to air temperature at 1.2 m above ground (and in minimum shade, 40%) between 2019-01-15 and 2019-01-31."
#| code-fold: true
  
  # Load the plant object
    load(here("output", "microclim", "plant_micro.RDa"))
    load(here("output", "microclim", "plant2_micro.RDa"))
  
  # Extract the microclimate temperatures. Becasue this was simulated for a plot the 'body tempertaure'is the leaf temperature and is stored as 'TC' in the environ object
            environ_plant_plantecophys2 <- as.data.frame(plant2$environ)
      environ_plant_plantecophys2$dates <- micro$dates  # For clarirty add in dates from microclimate that ecto model uses

  environ_plant_plantecophys <- environ_plant_plantecophys2 %>% filter(dates > "2019-01-15" & dates < "2019-01-31") %>%
              dplyr::select(dates, TC) %>%
              pivot_longer(cols = c(TC), 
                           names_to = "Temps", 
                           values_to = "Temperature") %>% 
              mutate(Temps = ifelse(Temps == "TC",  "Leaf Temperature (Stomatal Conductance)"))

  # Combine the two dataframes environ_plant_plantecophys and environ_plant2
  new_environ_plant2 <- rbind(environ_plant_plantecophys, environ_plant2)

  ggplot(new_environ_plant2) + 
              geom_line(aes(y = Temperature, x = dates, col = Temps), show.legend = TRUE) +
              labs(y = "Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
              scale_colour_viridis_d()
```

Heer we can see that stomatal closure in teh middle of the day causes the leaf temperatures to rise substantially compared to if they were always open and transpiring, which would result in evaporative cooling. This is important because it means that the grasshopper will be exposed to higher temperatures. 

### **Step 2: Cross-talk between plant and animal systems: Incorporating cooling effects of plant transpiration to estimate grasshopper body temperature** {#step2}

Now, lets run the ectotherm model for a grasshopper, but instead of using the micrioclimate object as it is, lets assume that the plant environment forms the mircoclimate of the grasshopper. This allows us to take in the microclimate and use it to calculate the body temperature of the grasshopper, and eventually uinderstand how this affects growth, development etc within a dynamic energy budget model. The ectotherm model also allows us to incorporate relevant behavioural plasticity in response to extreme heat, which we'll do. 

```{r, ecto_grasshopper}
#| label: ecto_grasshopper
#| echo: true
#| warning: false
#| message: false

# First, lets fix up the microclimate object so that we can use it in the ectotherm model. We need to add in the leaf temperature as the body temperature of the grasshopper. There are a couple of ways we can conceptualise this. For example, we could assume that animals are in shade and therefore infrared radiative heat (`TSKYC`) is the temperature the grasshopper experiences from their environment. Alternatively, we could assume that when sitting on a leaf the grasshopper is in direct contact with the leaf and therefore the leaf temperature is the body temperature. 
            
             # micro$metout[,"TC"] <- micro
             #      grasshopper$TC <- T_leaf
             #   grasshopper$dates <- micro$dates

# Parameters for the grasshopper
    Ww_g <- 0.2  # g, body mass
   shape <- 1    # cylinder
 shape_b <- 10   # ratio of long to short axis of cylinder
  T_pref <- 38   # preferred body temperature (animal will attempt to regulate as close to this value as possible)
 T_F_min <- 13   # degrees C, lower body temperature for foraging
 T_F_max <- 40   # degrees C, upper body temperature for foraging
 T_B_min <- 12.5 # degrees C, minimum basking temperature
T_RB_min <- 12.5 # degrees C, temperature at which animal will move to a basking site
  CT_max <- 47   # degrees C, critical thermal maximum (animal will die if CT_kill = 1 and this threshold is exceeded)
  CT_min <- 0    # degrees C, critical thermal minimum (used by program to determine depth selected when inactive and burrowing)


# Behavioural traits. This tells the model what the grasshopper can do in response to heat. 
     diurn <- 1     # diurnal activity allowed (1) or not (0)? 
   nocturn <- 0     # nocturnal activity allowed (1) or not (0)?
    crepus <- 1     # crepuscular activity allowed (1) or not (0)?
shade_seek <- 1     # shade seeking allowed (1) or not (0)?
    burrow <- 0     # shelter in burrow allowed (1) or not (0)?
  pct_cond <- 0     # Percentage of animal surface contacting the substrate

# Now that we have set the parameters we can run the model
grasshopper <- ectotherm(Ww_g = Ww_g,
                        shape = shape,
                      shape_b = shape_b,
                       T_pref = T_pref,
                      T_F_min = T_F_min,
                      T_F_max = T_F_max,
                      T_B_min = T_B_min,
                     T_RB_min = T_RB_min,
                       CT_min = CT_min,
                       CT_max = CT_max,
                        diurn = diurn,
                      nocturn = nocturn,
                       crepus = crepus,
                   shade_seek = shade_seek,
                       burrow = burrow,
                     pct_cond = pct_cond
)

save(grasshopper, file = here("output", "microclim", "grasshopper_micro.RDa"))
```

Now that we have run the ectotherm model we can plot the body temperature of the grasshopper over time along side the air temperature. This will allow us to see how the grasshopper responds to the heatwave.

```{r, fig-grasshopper_temp}
#| label: fig-grasshopper_temp
#| fig-cap: "Body temperature of a grasshopper at a location near Renmark, South Australia. Body temperature of the grasshopper is shown in relation to air temperature at 20 cm and 1.2 m above ground (and in minimum shade, 20%) for a short times between 2019-01-15 and 2019-01-31."

# Load grasshopper object
  load(here("output", "microclim", "grasshopper_micro.RDa"))

# Get the resulting heat budget calculations for the grasshopper
      grasshopper_environ <- as.data.frame(grasshopper$environ)
grasshopper_environ$dates <- micro$dates

# Filter to heat wave dates
grasshopper_environ2 <- grasshopper_environ                                                           %>% 
                        filter(dates > "2019-01-15" & dates < "2019-01-31")                           %>% 
                        select(dates, TA, TC, ACT)                                                    %>%
                        pivot_longer(     cols = c(TA, TC), 
                                      names_to = "Temps", 
                                     values_to = "Temperature")                                       %>% 
                        mutate(Temps = ifelse(Temps == "TA", "Air Temperature", "Body Temperature"))

  ggplot(grasshopper_environ2) + 
              geom_line(aes(y = Temperature, x = dates, col = Temps), show.legend = TRUE) + 
              geom_point(aes(y = ACT, x = dates, fill = factor(ACT)), shape = 21, size = 4) +
              labs(y = "Temperature (°C)", x = "Date") + theme_classic() +
              theme( axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
              scale_colour_viridis_d() +
              scale_fill_manual(values = c("0" = "blue", 
                                           "1" = "green", 
                                           "2" = "orange"),
                                labels = c("0" = "Inactive", 
                                           "1" = "Basking", 
                                           "2" = "Foraging")) +
              labs(fill = "Behaviour")
```

# **Using a systems modelling approach to help develop solutions to extreme heat events**

A systems modelling approach can also be used to design possible solutions by understanding better how 'thermal interventions' affect heat experienced by organisms and how it influences growth, survival and reproduction. For example, we can manipulate rainfall at a particular time to simulate artficial watering that might be a possible intervention used during extreme heat. The added water can be used to make calculations of heat exchange, temperatures and transpiration under heat events. We can also optimise the amount of time to balance costs and benefits of watering.

```{r, microclimate_water}
#| label: microclimate_water
#| echo: true
#| warning: false
#| message: false
#| eval: false

# Lets reload our microclimate object so we can use it again
  load(here("output", "microclim", "microclimate.RDa"))

# Lets now create a couple scenerios. This includes: 1) high shade conidtions with no water, 2) low shade conditions with no water, 3) high shade conditions with water and 4) low shade conditions with water.
    # First, get high and low shade conditions from our origial simulation
     micro_lowshade_nowater <- micro$metout
    micro_highshade_nowater <- micro$shadmet


RAINFALL <- micro_lowshade_nowater$RAINFALL # extract rainfall from previous sim
dates2 <- micro_highshade_orig$dates2
dates <- micro_highshade_orig$dates



plot(dates2, RAINFALL, type = "h")
date_waters <- as.character(seq(as.POSIXct("2019-01-01"), as.POSIXct("2019-01-31"),
by = "days"))

# Now lets add in the water. We can do this by adding a rainfall event to the microclimate object. This is done by creating a new vector of rainfall and then adding it to the microclimate object.
      water <- 10 # mm water to add
  RAINFALL2 <- RAINFALL
RAINFALL2[as.character(dates2) %in% date_waters] <- water
plot(dates2, RAINFALL2, type = "h")
# Create hourly values
rainhour <- unlist(lapply(RAINFALL2, function(x) c(x, rep(0,
23))))
maxshade <- 90 # maximum shade available
micro_highshade_water <- micro_silo(loc = loc, dstart = dstart,
dfinish = dfinish, Usrhyt = Usrhyt, microclima = microclima,
maxshade = maxshade, PC = PC, SP = SP, rainhourly = 1, rainhour = rainhour,
email = "m.kearney@unimelb.edu.au")
micro_highshade_orig <- micro_highshade_water # save a copy because will be modifiying it maxshade <- 5 # maximum shade available
micro_lowshade_water <- micro_silo(loc = loc, dstart = dstart,
dfinish = dfinish, Usrhyt = Usrhyt, microclima = 1, maxshade = maxshade,
PC = PC, SP = SP, rainhourly = 1, rainhour = rainhour, email = "m.kearney@unimelb.edu.au")
micro_lowshade_water_orig <- micro_lowshade_water # save a copy because will be modifiying it later
save(micro_highshade_water, file = "micro_highshade_water.Rda")
save(micro_lowshade_water, file = "micro_lowshade_water.Rda")

```