---
title: "A systems-modelling approach to predict biological responses to extreme heat"
date: "`r Sys.Date()`"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "index.html"
    theme: cosmo
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
---

## Introduction

In this supplementary material, we provide extended versions, with code, of our case studies we draw upon within our manuscript to demonstrate how a systems modelling approach can be implemented in practice. Our case studies focus on plant-insect-microbe interactions, and how these interactions are influenced by extreme heat events. We provide the code for the models we use in our case studies, and the code for the simulations we run to predict the impact of heatwaves on these interactions. It is important to realize that extreme heat events are relative to the system in question. What qualifies as 'extreme' will vary across species and communities. Models and code will of course always need to be modified to suite the specific system in question and should be validated against empirical data.

## Load packages

Our case studies all use the statistical software [R](https://www.r-project.org/). We use a number of packages to run our simulations, including [`NicheMapR`](https://mrke.github.io) [@Kearney2009b; @Kearney2013], [`microclima`](https://github.com/ilyamaclean/microclima) [@Maclean2019; @Kearney2020], and [`plantecophys`](https://github.com/RemkoDuursma/plantecophys) [@Duursma2015].  The code below loads the required packages that are needed for this tutorial.

```{r, libraries}
#| label: libraries
#| echo: true
#| warning: false
#| message: false

# You will need to install a few packages first, a couple from GitHub as they are not on CRAN. Note that you will need to remove '#' from the code below to install the packages.
# install.packages("pacman")
# devtools::install_github('mrke/NicheMapR')
# devtools::install_github("ilyamaclean/microclima")

# Load required libraries after the installation
pacman::p_load(devtools, deSolve, ggplot2, dplyr, tidyr, purrr, patchwork, equatags, flextable, NicheMapR, terra, microclima, plantecophys, magick, here)

```

:::{.column-margin}
:::{.callout-note}
gFortran can be a problem for NicheMapR so you need to set the paths properly if you are getting errors. Follow details [here](https://stackoverflow.com/questions/69639782/installing-gfortran-on-macbook-with-apple-m1-chip-for-use-in-r). The following code will help you set the paths properly.

```{r}
#| echo: true
#| eval: false
#| results: hide
#| code-fold: true

# But do the following:
# 1. Install gfortran using homebrew in the terminal/bash window
   brew install gcc 
# 2. In home directory of terminal create a Makevars file
   mkdir ~/.R
   touch ~/.R/Makevars
   open  ~/.R/Makevars
# 3. Add the following to the Makevars file. Note that you need to make sure your version of fortran is correct and replace 124.2.0_1 with your version. You can find version in terminal by typing `gfortran --version`
    FC = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
   F77 = /opt/homebrew/Cellar/gcc/14.2.0_1/bin/gfortran
 FLIBS = -L/opt/homebrew/Cellar/gcc/14.2.0_1/lib/gcc/11

#4. Restart R and try to install the package again. It should now work.
```
:::
:::

## Case Study 1: Simulating plant and insect interactions under heatwaves

Case study one explores how feedback between plants and insects can happen in response to extreme heat events. Here, we are interested in how the behaviour of grasshoppers is influenced by changes in plant transpiration which can impact microhabitat temperatures of grasshoppers (@fig-case1). While this is a simple situation it's a useful starting point to understand how a systems modelling approach can be used to predict the impact of heatwaves on species interactions. 

```{r}
#| label: fig-case1
#| echo: false
#| fig-cap: "Grasshopper and plant interactions under heatwaves. Grasshoppers feeding on leaves of a bush in Calperum Station, South Australia. Complex leaf structure provides shade and cooler microclimate for grasshoppers. Leaves exposed to hot temperatures close their stomata to conserve water reducing transpiration and photosynthesis. Leaves that are not exposed to direct sunlight can keep their stomata open and continue photosynthesis while also transpiring and providing cooler microhabitats that grasshoppers can exploit through thermoregulatory behaviours. Importantly, stomatal behaviour is mediated by drought conditions. More stomata will be closed during drought to conserve water and this will impact the cooling effect of the plant on the grasshopper."

fig_case1 <- image_read(here("figs", "figS1.png"))
fig_case1
```


### Step 1: Predicting the microclimate of a grasshopper

To predict the impact of heatwaves on plant-insect interactions, we first need to predict the microclimate. Using biophysical models we can use coarse climate, topological, rainfall, elevation and spatial information to make predictions on what the mircoclimate a species is likely to experience, which can help us predict it's likely body temperature (*exposure*). It is important to recognize that these are predictions. As with any model, they need field validation [@De_Frenne2024]. Nonetheless, they provide an important tool in our systems modelling framework to help us understand how heatwaves are likely to affect species interactions.

To model the microclimate, we can use [`NicheMapR`](https://mrke.github.io) and the [`microclima`](https://github.com/ilyamaclean/microclima) package to predict the microclimate at a given location. We can then use this microclimate data to simulate the growth of plants and insects under different temperature conditions [@fig-case1].

### Deb Modelling Paper

Now that we have `NicheMapR` lets try doing some DEB modelling for, say, a single location for a number of species.

```{r}
#| echo: true
#| eval: false
#| results: hide
#| code-fold: true
#| 
# First, set the location where NCEAP climate data is located. This is a global dataset. We need this to calculate the microclimate. Only need to do once.
   get.global.climate(folder="./NCEP_dan/NCEAP/")

# Location we want
  longlat <- c(140.79970255974192, -34.159176277782564) # Near Renmark
    micro <- micro_global(loc = longlat, nyears = 15, soiltype = 9, minshade = 40, maxshade = 80, Usrhyt = 0.2, terrain = 0)

# Explore microclimate data
  soil_temp <- micro$soil

ggplot(soil_temp, aes(y = D0cm, x = TIME, col = DOY)) + geom_line() + facet_wrap(~DOY)
```


## Case Study 1: Predicting the impact of heatwaves on the growth of plants, microbes and insects

First, we need to get the microclimate, or simulate our own with some heatwaves. This can be done as follows:

```{r, micro_sim1}

#| label: micro_sim1
#| echo: true
#| eval: false

# Set up environmental model (e.g., microclimate)# 30 days simulation
	#microclim <- microclimate(loc = c(30, -85),   month = 6, n = 30, shade = 0.5)  # 50% shade

# Generate temperature data for DEB models
#temperature <- microclim$output$temp 

# Set random seed for reproducibility
set.seed(42)

# Generate a base temperature time series (e.g., daily data for 1 year)
days <- 1:365  # 1 year of daily data
base_temp <- 20 + 10 * sin(2 * pi * days / 365)  # Seasonal temperature

# Add random daily variation
daily_variation <- rnorm(length(days), mean = 0, sd = 2)
temperature_series <- base_temp + daily_variation

# Add a few heatwave events (spikes in temperature)
heatwave_days <- c(50, 120, 200, 280)  # Arbitrary days for heatwaves
heatwave_durations <- c(5, 7, 4, 6)  # Duration of each heatwave
heatwave_intensity <- 8  # Degrees Celsius above the base temperature

for (i in seq_along(heatwave_days)) {
  start <- heatwave_days[i]
  duration <- heatwave_durations[i]
  temperature_series[start:(start + duration - 1)] <- temperature_series[start:(start + duration - 1)] + heatwave_intensity
}

# Create a data frame for the time series
temperature_df <- data.frame(
  Day = days,
  Temperature = temperature_series
)

# Plot the temperature time series
library(ggplot2)
ggplot(temperature_df, aes(x = Day, y = Temperature)) +
  geom_line(color = "blue") +
  geom_point(data = data.frame(Day = heatwave_days, 
                               Temperature = temperature_series[heatwave_days]), 
             aes(x = Day, y = Temperature), color = "red", size = 3) +
  labs(title = "Simulated Temperature Time Series with Heatwaves",
       x = "Day of Year", 
       y = "Temperature (°C)") +
  theme_minimal()
```

Now that we have the temperatures each species is occupying we can include this temperature time series in our DEB models. We can simulate the growth of plants, microbes and insects under these conditions. The outputs of each can be used as input for others. First, we need to create a dynamic energy budget model for each of the three interacting species.

```{r, deb_sim1_function}
#| label: deb_sim1_function
#| echo: true
#| eval: false

# Define DEB model for insect pest
insect_deb <- function(temp, food) {
  kappa <- 0.8  # Fraction of energy allocated to maintenance and growth
  growth <- kappa * food * exp(-((temp - 25)^2) / 10)  # Simplified growth equation
  reproduction <- (1 - kappa) * food * exp(-((temp - 30)^2) / 15)
  list(growth = growth, reproduction = reproduction)
}

# Define DEB model for plant
plant_deb <- function(temp, light, nutrient) {
  photosynthesis <- light * exp(-((temp - 20)^2) / 10)  # Simplified photosynthesis
  growth <- photosynthesis * nutrient
  respiration <- 0.1 * photosynthesis
  list(growth = growth, respiration = respiration)
}

# Define DEB model for microbe
microbe_deb <- function(temp, organic_matter) {
  growth <- organic_matter * exp(-((temp - 30)^2) / 10)
  respiration <- 0.2 * growth
  list(growth = growth, respiration = respiration)
}
```

Now, we can run the simulation for each species.

```{r, deb_sim1}
#| label: deb_sim1
#| echo: true
#| eval: false

# Simulate interactions over 30 days
days <- 1:30
results <- data.frame(day = days, 
                      insect_growth = NA, 
                      plant_growth = NA, 
                      microbe_growth = NA)

# Initialize resources
plant_nutrients <- 1
microbe_organic_matter <- 1
light <- 1  # Assume constant light for simplicity

for (day in days) {
  temp <- temperature[day]
  
  # Insect feeds on plant
  insect_result <- insect_deb(temp, food = plant_nutrients * 0.1)
  
  # Plant photosynthesis and growth
  plant_result <- plant_deb(temp, light, nutrient = plant_nutrients)
  
  # Microbe decompose organic matter
  microbe_result <- microbe_deb(temp, organic_matter = microbe_organic_matter)
  
  # Update resources
  plant_nutrients <- plant_nutrients + plant_result$growth - insect_result$growth
  microbe_organic_matter <- microbe_organic_matter + insect_result$reproduction - microbe_result$respiration
  
  # Record results
  results$insect_growth[day] <- insect_result$growth
  results$plant_growth[day] <- plant_result$growth
  results$microbe_growth[day] <- microbe_result$growth
}

```

Now that we have the results, we can plot them to see how the growth of each species changes over time.

```{r, deb_sim1_plot}
#| label: deb_sim1_plot
#| fig-cap: "Growth of plant, insect and microbe over 30 days"
#| fig-pos: H
#| echo: true
#| eval: false

# Plot results
library(ggplot2)
ggplot(results, aes(day)) +
  geom_line(aes(y = insect_growth, color = "Insect Growth")) +
  geom_line(aes(y = plant_growth, color = "Plant Growth")) +
  geom_line(aes(y = microbe_growth, color = "Microbe Growth")) +
  labs(y = "Growth Rate", color = "Species") +
  theme_minimal()
```

## Case Study 2: Developing a DEB model for parasitoid wasps applied to control of an aphid system under heatwaves

In this case study, we will develop a DEB model for a parasitoid wasp that preys on aphids. We will then simulate the growth of the aphid population and the parasitoid wasp population under different temperature conditions, including heatwaves. Following from @Llandres2015, we will use the DEB model to predict the impact of heatwaves on the parasitoid wasp population and the aphid population. @Llandres2015 build a multi-life stage model for the wasp. 


#### Building a DEB model for the parasitoid wasp

To begin, we need to first define the DEB model for the various life-stages. Lets first define the key metabolic processes as functions:


$$
\dot{v}
$$

```{r, tbl-equivalent}
#| label: tbl-equivalent
#| tbl-cap: "Mathematical notation and equivalent code notation"

eqn <- c(
"\\dot{p}_{A}",
"[\\dot{p}_{Am}]",
"f",
"V",
"\\dot{p}_{C}",
"\\dot{v}",
"E",
"[E_{G}]",
"[\\dot{p}_{S}]", 
"\\dot{k}_{E}",
"\\kappa",
"[E]",
"\\dot{p}_{S}",
"[\\dot{p}_{M}]",
"\\dot{p}_{G}",
"\\dot{p}_{J}",
"\\dot{k}_{J}",
"\\dot{p}_{R}",
"E_{H}^b"
)

code <- c(
"pA",
"pAm", 
"f",
"V",
"pC",
"v",
"E",
"Eg_brac",
"pS_brac", 
"kE",
"kappa",
"E_brac",
"pS",
"pM_brac",
"pG",
"pJ",
"kJ",
"pR",
"Ehb"
)

# Create dataframe
df <- data.frame(formula = eqn, code = code)

# Create table
ft <- flextable(df)
ft <- compose(
    x = ft, j = "formula",
    value = as_paragraph(as_equation(formula, width = 2, height = .5))
  )
ft <- flextable::align(ft, align = "center", part = "all")
ft <- flextable::width(ft, width = 2)
ft
```


```{r, deb_paramaters}
#| label: tbl-deb_paramaters
#| tbl-cap: "Parameters for the DEB model"
#| echo: true

# Define parameters for the DEB model
parameter  <- c("kx", "f", "Eg", "v_dot", "pM", "k", "kJ", "kR", "z", "DeltaM", "ha", "sG", "yEV", "sj", "sS", "Exh", "Ebh", "Eeh", "Eeggsh", "kE", "kSA", "kSG", "kSD")

detail <- c("assimilation efficiency","Holling type II scaled functional response","volume-specific cost of structure","energy conductance","volume-specific somatic maintenance","fraction of mobilized reserve allocated to soma","maturity maintenance rate coefficient","fraction of the reproduction buffer fixed into eggs","zoom factor","shape correction coefficient","Weibull ageing acceleration","Gompertz stress coefficient","yield of imago reserve on larval structure","threshold for the onset of pupation","threshold for the onset of silk production","maturity threshold from embryo1 to embryo2","maturity threshold from the embryonic stage to the larval stage maturity threshold at imago" ,"emergence","maturity threshold at egg maturation","reserve and reproduction buffer turnover of the imago","mass coefficient from assimilation to silk production","mass coefficient from growth to silk production","mass coefficient from dissipation to silk production")

values  <- c(0.94, 0.95, 21210, 0.0207, 245.7, 0.930, 0.002, 0.85, 3.98, 0.152, 0.004, 0.0001, 0.8, 0.983, 0.106, 0.0002, 0.044, 5.56, 5.48, 0.291, 0.03, 0.5, 0.94)

flextable::flextable(data.frame(parameter, detail, values))  %>% flextable::set_table_properties(width = .5) %>% flextable::set_table_properties(layout = "autofit") %>% flextable::theme_zebra() %>% flextable::theme_booktabs()
```


```{r, deb_model}
#| label: deb_model 
#| echo: true
#| eval: false

# At any given time we can calculate the key metabolic processes. These then impact the state variables, Reserve, Structure, Maturation (Eh) and Reproduction (Er).

####----------------------------------------------#######
# Larval Stage
####----------------------------------------------#######
  
  pA <- function(f, V, pAm){
    return(pAm * f * V) # Table 5 from Llandres2015
  }

  pC <- function(E, Eg, V, pS_brac, kE, kappa, E_brac){
    return(E * ((((Eg_brac * kE) + pS)/(kappa*E_brac + Eg))) # Table 5 from Llandres2015
  }

  pS  <- function(pM_brac, V){
    return(pM_brac*V) #Table 5 from Llandres2015
  }

  pG  <- function(pC, pM, kappa){
    return(kappa*pC - pM) #Table 5 from Llandres2015
  }

  pJ  <- function(kJ, Ehb){
    return(kJ * Ehb) #Table 5 from Llandres2015
  }

  pR  <- function(kappa){
    return((1 - kappa)*pC - pJ) #Table 5 from Llandres2015
  }

####----------------------------------------------#######
# Pupal Stage
####----------------------------------------------#######

pL_transformation_pupae  <- function(V, kEl){
    return(V*kEl) #Table 5 from Llandres2015
}

pc_mobilisation_E_pupae  <- function(E, vj, Eg, V2, pS, ke, k){
    return(E * (((((Eg * vj) / (V2^1/3)) + pS)/(k*E + Eg)))
}

pS_somatic_maintenance_pupae  <- function(pM, V2){
    return(pM*V2) #Table 5 from Llandres2015
}

pG_growth_pupae  <- function(pC, pM, k){
    return(k*pC - pM) #Table 5 from Llandres2015
}

pJ_maturity_maintenance_pupae  <- function(kj, Eh){
    return(kj * eh) #Table 5 from Llandres2015
}

pR_maturation_pupae  <- function(k){
    return((1 - k)*pC - pJ) #Table 5 from Llandres2015
}

pc2_mobilisation_Er_pupae  <- function(ke, Er){
    return(ke*Er) #Table 5 from Llandres2015  
}

pro_repro_overhead_pupae <- function(kr, pc2){
    return((1-kr)*pc2) #Table 5 from Llandres2015
}

pr2_eggFlux_pupae <- function(pc2, pro){
    return(pc2 - pro) #Table 5 from Llandres2015
}

####----------------------------------------------#######
# Imago Stage
####----------------------------------------------#######

pa_assimilation_imago <- function(ps, pj){
    return(ps + pj) #Table 5 from Llandres2015
}

pc_mobilisation_E_imago <- function(ke, E){
    return(ke*E) #Table 5 from Llandres2015
}

ps_somatic_maintenance_imago <- function(pM, V2){
    return(pM*V2) #Table 5 from Llandres2015
}

pj_maturity_maintenance_imago <- function(kj, Ehe){
    return(kj * Ehe) #Table 5 from Llandres2015
}

pr_repro_imago <- function(pc, ps, pj){
    return(pc - ps - pj) #Table 5 from Llandres2015
}

pc2_mobilisation_Er_imago <- function(ke, Er){
    return(ke*Er) #Table 5 from Llandres2015
}

pro_repro_overhead_imago <- function(kr, pc2){
    return((1-kr)*pc2) #Table 5 from Llandres2015
}

pr2_eggFlux_imago <- function(pc2, pro){
    return(pc2 - pro) #Table 5 from Llandres2015
}
```

@tbl-deb_paramaters shows the parameters for the DEB model. These parameters are used in the functions defined above to calculate the metabolic processes.

```{r, simulations}
#| label: simulations
#| echo: true
#| message: false
#| warning: false
#| eval: false

# Lets just go through the pupal stage for now
# Parameters
  E  =
  vj = 
  Eg = 
  V2 = 
  pS = 
  ke = 
  k  = 
  f  = 
  V  = 
  pAm= 
  pM = 
  V  =
  pT =


```


