---
title: "Thermal tolerance landscapes across populations and communities: applications in a changing world"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---
# Introduction

Climate change is expected to result in warmer and more variable thermal environments globally resulting in more frequent and severe heat wave events – events that result in prolongered periods of temperatures supassing the 90th percentile of historical temperatures [@barriopedro2023heat; @aglas2024recent]. Greater thermal variability will result in organisms experiencing more extreme heat wave events that exceed species tolerance limits. Exceeding physiological limits will have profound impacts on orgamismal growth, survival and fitness with cascading effects on population dynamics, community composition and ecosystem services that human populations rely upon. Climate variability has already resulted in dramatic declines in pollinator abundance [@kazenel2024heat] and in crop yields globally [@zhao2017temperature; @tito2018global], but the causes underlying such declines are not well understood. Declines are likely driven by a combination of direct effects of heat stress on plant physiology and indirect effects on plant interactions with other species, including the microbial and insect communities with which they interact [@trivedi2022plant]. Nonetheless, understanding the dynamic interplay between direct and indirect interactions has been hampered by the complexity of community interactions, lack of methods for measuring and predicting changes in interaction strength in diverse communities and how such interactions impact community assembly [@trivedi2022plant]. Understanding how thermal tolerance varies across populations and communities is crucial for predicting how climate change will affect ecosystems and the services they provide.


Márquez *et. al* @marquez2007virus describe a three-way mutulastic assocation between a tropical panic grass (*Dichanthelium lanuginosum*), the fungus, *Curvularia protuberata*, and a *Curvularia* virus whereby thermal tolerance of the fungus and grass is highly dependent on fungal-virus associations. Endophytic fungus cured of the virus were unable to survive extreme heating events (temperatures of 65$^\circ$C for 2 weeks). However, when the fungus were inoculated with the virus, heat tolerance was restored. Fascinatingly, the virus could be manipulated to confer heat tolerance to tomato (*Solanum lycopersicon*) demonstrating the potential for conferring thermal tolerance to other plant species. The mechanisms by which fungal-virus associations confer thermal tolerance are not yet understood but are hypothesized to involve scavenging of reactive oxygen species (ROS) by fungal endophytes generated by plant defense mechanisms in response to heat stress. 

Li *et al* @li2023interspecific measured the thermal tolerance landscape for three aphid species that are pests for cereal crops globally to understand whether it could predict changes in community abundance. Using a thermal death time (TDT) approach to estimate the culmuative impacts of heat stress on survival they found that one species, *Rhopalosiphum padi*, showed high heat tolerance and accumulated less injury over natural temperature fluctuations, which may explain the increase in relative abundance of this species with extremely high temperatures observed in southern parts of its range. Li *et al* @li2023interspecific also found that different life-stages were affected differently by heat stress, with adults being more thermally tolerant than early developmental stages emphasising the need for models to capture such variation. However, such findings do not incorperate acclimation to heat stress or different microhabitat choice by the different species, which are both likely to be important for predicting the impacts of heat stress on community structure. It also ignores how interspecific interactions between competing aphid species will also likely mediate species abundance. It would be worth while integrating thermal tolerance landscapes with ecological models that capture interspecific interactions to better predict how heat stress will affect community structure.

# Modelling thermal tolerance landscapes to capture interactions*

-chat with Marggie and Mike about this some more. But see @zou2024time

# Methods for manipulating thermal tolerance to maintain functional interactions

*Genetic engineering*: The use of genetic engineering to confer thermal tolerance to plants and animals is a promising avenue for enhancing thermal tolerance. For example, the overexpression of heat shock proteins (HSPs) in plants has been shown to enhance thermal tolerance. HSPs are a family of proteins that are upregulated in response to heat stress and are involved in protein folding and degradation, and the prevention of protein aggregation. CRISPR-Cas9 gene editing technology has been used to enhance thermal tolerance in plants by modifying genes involved in heat stress response.

*Selective breeding*: Artifical selection will be a useful avenue for both plants and animals to enhance thermal tolerance. 

# Applications of thermal tolerance landscapes in a changing world

## Differenecs in focus for plants and animals
- Animals behave. We need to model that. HOw do changes in behaviour impact inetractions with other species and the environment. How might this impact plant productivity   

## Network theory can integrate thermal tolerance landscapes with ecological models
- seems like most people are analysing communities with network theory. Some important papers @delmas2019analysing @godoy2019coexistence @poisot2021global @hui2019invade @gomez2023role @lytle2023matrix @petsopoulos2021using Should also look at this book @dale2021quantitative

have a close look at @vitali2022disruption and @riva2023toward

limitations @terry2020finding

Also this is a cool paper @lu2023food on carnivores diets by DNA sequecing fecal data

The other way to model interactions us using foundational communitu ecology/ species abundance models see @carrara2015inferring. We can here model the actual abundance of a given species and it would seem more flexible to incorperate thermal tolerance infor into these models. 

# Future Directions

*Better understanding of the mechanisms enhancing thermal tolerance in plants and animals*: We need to understand the genomic, cellular and physiological processes that confer greater thermal tolerance and help prepare organisms for future heat stress events.

*Modelling inetractions among plants and animals to build community thermal tolerance landscapes and predict how heat stress events will affect communities*: We need to understand how thermal tolerance of individual species will affect the thermal tolerance of communities and ecosystems.

*Trade-offs between thermal tolerance and other traits*: We need to understand the trade-offs between thermal tolerance and other traits such as growth rate, reproduction, and competitive ability. It is fine to confer greater thermal tolerance resistance but if this comes at a cost to reproduction, growth or biomass accumulation, then it may not be beneficial. NEED EXAPLES

# References


```{r, performancecurve}
## See box 1 in Weiss-Lehman et al. 2022 for a decsription of the model. This could be applied to non-linear fitness effects based on optimal environment.

#parameters
bmax = 2  # Maximum pop growth rate lamda (think of this as optimal fitness)
 zi  = 5  # Optimal environment (the optimal temperature where fitness is maximised)
 sdi = 3  # Niche breadth (this is the variation in the fitness surface)
  xe = seq(-10,10,0.1) # Environmental gradient (self-explanatory)

lamda <- bmax*exp(-((zi-xe)/(2*sdi))^2)
plot(xe,lamda, type = "l", ylab = "Population growth rate (Lambda)", xlab = "Environmental gradient")

## Another way of doing this from Agelettas book pg 229
#parameters
 rm = 2  # Maximum pop growth rate lamda (think of this as optimal fitness)
 z  = 5  # Optimal environment (the optimal temperature where fitness is maximised)
 sd = 10  # Niche breadth (this is the variation in the fitness surface)
 tb = seq(-10,10,0.1) # Environmental gradient (self-explanatory)
r <- rm - ((z-tb)^2/(2*sd^2))

plot(tb,r, type = "l", ylab = "Fitness", xlab = "Environmental gradient")


```

```{r, skewnormal}
# Skew normal fitness functins from https://academic.oup.com/evolut/article/73/2/145/6882140
#The parameter θ specifies the location, sigma the width, and alpha the skew of the fitness function. α > 0 gives a right-skewed function (positive skewness), α < 0 a left-skewed function, and α = 0 a symmetric Gaussian fitness function. Importantly, α also changes the position of the maximum value of w (despite location parameter θ being kept constant), so we will write θ0 as the value of θ that provides maximum fitness for a trait value of zero.
#parameters
delta <- 5  # Optimal environment (the optimal temperature where fitness is maximised)
sigma <- 10  # Niche breadth (this is the variation in the fitness surface)
z = seq(-10,15,0.1)
alpha = -1.2
w <- 2*exp(-(z-delta)^2/(2*sigma^2))*pnorm(alpha*(z-delta)/sigma)

plot(z,w, type = "l", ylab = "Fitness", xlab = "Environmental gradient")

```
