---
title: "Predicting the resilience of species interactions under extreme heat"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: false
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

**Alternative Titles**: (just brainstorming)
- Integrated modeling approaches for predicting the impacts of extreme heat on communities
- Thermal tolerance landscapes for populations and communities: applications in a changing world
- An integrated modelling approach for predicting the impacts of extreme heat on communities
- Predicting the resilience of species interactions under extreme heat now and into the future
- Calculated responses to the biological impacts of extreme heat: the need to connect our models (Mike) (Dan likes this one as well) 
- Connecting our models for making calculated biological responses to extreme heat 

**Target Journal**:
*Trends in Ecology and Evolution*, Opinion
https://www.cell.com/trends/ecology-evolution/authors

Abstract: 100-120 words\
Highlights: Provide a short collection of 3-5 statements that concisely convey to the reader the recent advances in the field, including emerging concepts and/or distinctions, that constitute a main motivation for the discussion developed in the article. This summary is prominently featured on the first page of the article, thus allowing readers to immediately get a sense of the current status of the field.\
Word Count: 3500 words (this limit does not include text in the boxes, glossary, abstract, references, tables or figure legends)\
References: 80-100\
Outstanding Questions: 350 words\
Glossary: 500 words\

**Alternative homes**:

*Nature Ecology and Evolution* (Perspective) - 3000-4000 words, 4-6 display items: https://www.nature.com/natecolevol/content\
*Global Change Biology* - 3000 (Mini Review)-5000 (Opinion) words (https://onlinelibrary.wiley.com/page/journal/13652486/homepage/forauthors.html)\
*Ecology Letters* (Perspective) - up to 7500 words if needed. 

**Structure Overview - Sections**:\
- Abstract (Dan) (~120 words - First Draft Done)\
- A need for a mechanistically informed modelling framework for predicting the effects of extreme heat on community assemblages (Dan) (~ 500 words - First Draft Done)\
- The next wave of climate models\ (Mike, Belinda)

- Capturing the balance between physiological *damage* and *repair* to predict multi-species thermal sensitivity (Dan, Adrienne) (~300-500 words)\
- Biophysics to the rescue: realistic community wide *exposure* to extreme heat and Dynamic Energy Budget models (Mike, Belinda) (~300-500 word)\
- Mechanistic approaches that 'speak' to population and community ecology under extreme heat (Margie) (~300-500 words)\
- Incorporating context-dependent environmental feedback to evaluate resilience to extreme heat (Brajesh, Brendan, James, Everyone else who wants to contribute) (~500 words)\
- 'Hot' solutions for a changing world (Alison, Owen and Belinda, Mike, Anyone else!) (~500 words)\
- Future directions: modelling challenges, gaps and possible solutions (Mike, Belinda, Dan, Ary, Braj, Everyone) (~300-500 words)

Please see more detailed guidance for each section below and some names who might lead these.

# Abstract (Dan)
Anthropogenic climate change is leading to more frequent and extreme heat waves. These short-term but large-scale atmospheric events are radically re-shaping interactions among organisms, impacting biodiversity, community composition and ecosystem services crucial to natural systems and food security. Predicting heat wave impacts on interacting species requires an understanding of the processes driving differential exposure and sensitivity of organisms to extreme heat events in a life-cycle context. But this requires a modelling framework that can capture species-specific responses via habitat-mediated microclimates, biophysics of body temperature, physiological sensitivity, and its effects on vital rates across ontogeny and thermal history. We show how existing models in disparate fields can be linked to achieve this level of understanding to allow calculated responses to extreme heat now and into the future. 

# A need for a mechanistically informed modelling framework for predicting the effects of extreme heat on community assemblages (Dan)

Climate change is leading to warmer and more variable thermal environments globally [@barriopedro2023heat; @aglas2024recent]. Greater thermal variability will result in organisms experiencing extreme heat waves that result in thermal stress that impacts organismal growth, survival and reproduction with cascading effects on population dynamics, community composition and ecosystem structure and function. Climate variability has already been linked to dramatic declines in pollinator abundance [@kazenel2024heat] and in crop yields globally [@zhao2017temperature; @tito2018global], but the causes underlying such declines, and their ramifications through communities, are not fully understood. The impacts of extreme heat on species and communities are driven by a combination of direct effects of heat stress on organism physiology and fitness [@hesketh2023extreme; @martinez2024ecological; @rezende2020predicting; @orsted2024thermal] and indirect effects on interactions (both positive and negative) among species [e.g., @lopez2024heat; @huỳnh2024heatwave] -- including the microbial communities many species rely upon [@trivedi2022plant]. Understanding the dynamic interplay between direct and indirect interactions, and how these are mediated by environmental factors, has been hampered by the complexity of community interactions and the lack of an integrated modelling framework that can translate short-term physiological damage into changes in fitness and population growth across species [@trivedi2022plant].

Predicting how extreme heat impacts population dynamics and community interactions requires the coupling of physiological, biophysical and population ecology models. We argue that such a feat is becoming increasingly more tangible and will provide powerful new ways of predicting how extreme heat shapes community dynamics. Dynamic Energy Budget (DEB) theory can now be seamlessly integrated with biophysical models of heat and mass exchange to characterise organismal body temperatures and simulate the effects of extreme heat events on the entire life cycle of species within communities [@kooijman2010dynamic; @kearney2009mechanistic; @jusup2024untapped; @kearney2024fast; @briscoe2023mechanistic]. Dynamic Energy Budgets that incorporate estimates of thermal sensitivity across species can capture the delicate balance between damage and repair in physiological systems [@orsted2024thermal; @rezende2020predicting; @rezende2014tolerance] to translate the immediate and cumulative impacts of extreme heat on growth, survival and reproduction. Importantly, mechanistic physiological models provide common currencies that can then be integrated into population and community ecology models to capture how extreme heat events perturb interactions among species [**Margie**, some refs here you think would be good?].

Here, we bring together modelling approaches across diverse fields to establish a bridge for thinking about how we might begin integrating them to predict how extreme heat will affect communities. We focus our discussion on plant, insect and microbial communities given the strong interconnections between them, the ease with which they can be studied and their significance to natural and agricultural productivity. Nonetheless, our framework can be extended to incorporate other, diverse interactions.  We highlight the immense opportunities such an integration may provide through a greater understanding of the ways in which extreme heat stress manifests across species and by elucidating more clearly how and why interactions among species can change with implications for natural and agricultural systems, food security and ecosystem services. We discuss the challenges of applying such an integrated modelling framework moving forward and how it can be extended to include other critical features that capture evolutionary change and plasticity in response to extreme heat. 

# Capturing the balance between physiological *damage* and *repair* to predict multi-species thermal sensitivity (Dan, Adrienne)
- Introduce thermal load sensitivity theory
  - Heat effects depend on the *exposure* and *sensitivity* 
  - Mechanistically involves the accumulation of damage and repair over time in response to a temperature stressor
  - We know different species will be exposed to extreme heat at different life stages
  - We know sensitivity will vary across species and life stages
  - We can make predictions about impacts of heat on: 1) survival and 2) reproduction across life stages and even tissues
  - Can be applied to both plants and animals- allows for lagged effects of heat stress on survival and reproduction in response to short term heat stress events
  - Captured by modelling thermal effects of the severity of temperature and the amount of time an organism is exposed to that temperature on survival and reproduction of given life stages
  - Concepts can be applied to both plants and animals, and targeted to key tissues
  - Critical to TLS is the idea of damage accumulation and repair over time in response to a temperature stressor
  - End discussing the need to have realistic exposure to extreme temperatures. links to next section
  - Figure 1: demonstrating how thermal load sensitivity can be used to predict heat stress impacts on plants and animals, survival, reproductive tissue. Simplified. 

# Biophysics to the rescue: realistic community wide *exposure* to extreme heat and Dynamic Energy Budget models (Mike, Belinda, Dan)
- Combining TLS with biophysical and dynamic energy budget models 
- Introduce biophysical and dynamic energy budget models
  - Biophysical models are used to predict how temperature affects the energy balance of an organism
  - DEB models are used to predict how temperature affects the energy budget of an organism
  - Both models can be used to predict how temperature affects the growth, reproduction and survival of an organism
  - Combining these models with TLS can help predict how temperature affects the survival and reproduction of a focal species in a community

# Mechanistic approaches that 'speak' to population and community ecology under extreme heat (Margie) 
- How can we link mechanistic physiological models with population and community ecology theory?
- Introducing population and network models to link outputs from DEB (survival, reproduction, abundance) to population growth.
- How people are analysing communities with network theory. Some important papers @delmas2019analysing @godoy2019coexistence @poisot2021global @hui2019invade @gomez2023role @lytle2023matrix @petsopoulos2021using Should also look at this book @dale2021quantitative
- How can we capture interactions among species within the community to predict how heat stress will affect community structure? 
- Figure 2: demonstrating how TLS, biophysical and DEB models can be used to predict heat stress impacts on plants and animals, growth, reproduction, survival. (Dan, Margie)
- Also this is a cool paper @lu2023food on carnivores diets by DNA sequencing fecal data
-The other way to model interactions us using foundational community ecology/ species abundance models see @carrara2015inferring. We can here model the actual abundance of a given species and it would seem more flexible to incorporate thermal tolerance infor into these models. 
- have a close look at @vitali2022disruption and @riva2023toward, limitations @terry2020finding

# Incorporating context-dependent environmental feedback to evaluate resilience to extreme heat  (Brajesh, Brendan, James, Everyone else who wants to contribute)
- Goal of this section is to briefly focus on the BIOLOGY that will shift thermal exposure and sensitivity in plants and animals.
- I **just** found this important paper for which a lot can be gleaned upon. Many of the important things we have been discussing are discussed in here: @martinez2024ecological, so we should be careful to make our narrative distinct. The limitation of this study is it lacks the focus on modelling frameworks, so I think us distinguishing how we link these concepts back into the model would be quite novel and distinguish ourselves from them..
- Then, we overlay briefly how we might capture that in the modelling framework we setup above. For example, you all write the biology and we then need to think about how we can adjust models to capture these nuances. 
- If uncertain about adding a sentence to the link to the models above, that's ok, Dan, Mike, Margie and Belinda can add.

### Microbial communities can impact thermal sensitivity in plants and animals (Brajesh)

- Here's some text of a cool example I had written up. Feel free to use Braj or shift completely if you'd like. 
Interactions with symbiotic microbiota can also directly impact thermal tolerance of a focal species. For example, in plants, Márquez *et. al* @marquez2007virus describe a three-way mutualistic association between a tropical panic grass (*Dichanthelium lanuginosum*), the fungus, *Curvularia protuberata*, and a *Curvularia* virus whereby thermal tolerance of the fungus and grass is highly dependent on fungal-virus associations. Endophytic fungus cured of the virus were unable to survive extreme heating events (temperatures of 65$^\circ$C for 2 weeks). However, when the fungus were inoculated with the virus, heat tolerance was restored. Fascinatingly, the virus could be manipulated to confer heat tolerance to tomato (*Solanum lycopersicon*) demonstrating the potential for conferring thermal tolerance to other plant species. The mechanisms by which fungal-virus associations confer thermal tolerance are not yet understood but are hypothesized to involve scavenging of reactive oxygen species (ROS) by fungal endophytes generated by plant defense mechanisms in response to heat stress. 
- See @zou2024time. The paper provides a framework for modelling soil plant interactions

### Water availability can impact thermal sensitivity of plants and animals (Brendan)
### Competition can impact thermal sensitivity of plants and animals (James? Feel free to add a different section title here.)
- James. Added some text around a few examples below. Feel fre eto use/integrate or completely throw away. Or, we can move some of this into other sections.
Competition among species within a community can influence a focal species' tolerance to heat stress. Competition can influence microhabitat selection and the availability of thermal refugia which can protect against thermal extremes, the diet of a focal species that can indirectly influence physiological tolerance of temperature stress, and the availability of resources that can help buffer against heat stress. For example, Li *et al.* @li2023interspecific measured the thermal tolerance landscape for three aphid species that are pests for cereal crops globally to understand whether they could predict changes in community abundance. Using a thermal death time (TDT) approach to estimate the cumulative impacts of heat stress on survival they found that one species, *Rhopalosiphum padi*, showed high heat tolerance and accumulated less injury over natural temperature fluctuations. The ability to maintain high relative abundance in the face of extreme heat may allow this species to outcompete other aphids explaining why it dominates in southern parts of its range.

Li *et al.* @li2023interspecific also found that different life-stages were affected differently by heat stress, with adults being more thermally tolerant than early developmental stages emphasising the need for models to capture such variation. However, such findings do not incorporate acclimation to heat stress or different microhabitat choice by the different species, which are both likely to be important for predicting the impacts of heat stress on community structure. It also ignores how interspecific interactions between competing aphid species will also likely mediate species abundance. It would be worth while integrating thermal tolerance landscapes with ecological models that capture interspecific interactions to better predict how heat stress will affect community structure.

### Acclimation can impact thermal sensitivity of plants and animals (Everyone)

### Phenology and activity pattern shifts mediating interactions (James?)

# 'Hot' solutions for a changing world (Alison, Owen and Belinda, Mike)
- Here, we bring in what we can do with these models, I think we highlight some of the key elements of the solutions strand ideas where we can apply the models in the agricultural and natural space.
- Enhancing resilience
  -*Selective breeding*: Artificial selection will be a useful avenue for both plants and animals to enhance thermal tolerance. Models can help evaluate how changes in sensitivity can cascade to influence community structure and function (Alison, Owen, Dani, Others)
  - Microbial recruitment (Braj)
  - Pollinator services (James, Margie)
- Cooler landscapes
  - *Cooler landscapes*: Landscapes can be designed to provide thermal refugia for plants and animals. For example, planting trees in urban areas can provide shade and reduce the urban heat island effect. Similarly, planting hedgerows in agricultural landscapes can provide thermal refugia for beneficial insects and pollinators. (Mike, Belinda)
- Pest and disease prediction and control. (Ary)
- Conservation (Adrienne?)

# Future directions: modelling challenges, gaps and possible solutions

#### *Data Limitations for Applying DEB Theory*: (Mike) Requires lots of information on a given species to model life-cycles, challenges in applying stage-structured populations (metamorphosis) others....Solutions?

#### *Better understanding of the mechanisms enhancing thermal tolerance in plants and animals*: We need to understand the genomic, cellular and physiological processes that confer greater thermal tolerance and help prepare organisms for future heat stress events.

#### *Behavioural responses to heat stress*: Animals behave. We need to model that. How do changes in behaviour impact interactions with other species and the environment. How might this impact plant productivity  (Mike, Dan, DEB models)

#### *Evolutionary dynamics leading to changes in thermal sensitivity and exposure*: We need to understand how thermal sensitivity and exposure will evolve in response to climate change. Probably could cite @kearney2009integrating (Ary)

#### *Trade-offs between thermal tolerance and other traits*: We need to understand the trade-offs between thermal tolerance and other traits such as growth rate, reproduction, and competitive ability. It is fine to confer greater thermal tolerance resistance but if this comes at a cost to reproduction, growth or biomass accumulation, then it may not be beneficial. (Dani, James, Brendan)


# References

```{r, performancecurve}

## See box 1 in Weiss-Lehman et al. 2022 for a description of the model. This could be applied to non-linear fitness effects based on optimal environment.

#parameters
bmax = 2  # Maximum pop growth rate lamda (think of this as optimal fitness)
 zi  = 5  # Optimal environment (the optimal temperature where fitness is maximised)
 sdi = 3  # Niche breadth (this is the variation in the fitness surface)
  xe = seq(-10,10,0.1) # Environmental gradient (self-explanatory)

lamda <- bmax*exp(-((zi-xe)/(2*sdi))^2)
plot(xe,lamda, type = "l", ylab = "Population growth rate (Lambda)", xlab = "Environmental gradient")

## Another way of doing this from Agelettas book pg 229
#parameters
 rm = 1.5  # Maximum pop growth rate lamda (think of this as optimal fitness)
 z  = 35  # Optimal environment (the optimal temperature where fitness is maximised)
 sd = 8  # Niche breadth (this is the variation in the fitness surface)
 tb = seq(0,80,0.1) # Environmental gradient (self-explanatory)
r <- rm - ((z-tb)^2/(2*sd^2))

plot(tb,r, type = "l", ylab = "Fitness", xlab = "Environmental gradient")


```

```{r, skewnormal}
# Skew normal fitness functions from https://academic.oup.com/evolut/article/73/2/145/6882140
#The parameter θ specifies the location, sigma the width, and alpha the skew of the fitness function. α > 0 gives a right-skewed function (positive skewness), α < 0 a left-skewed function, and α = 0 a symmetric Gaussian fitness function. Importantly, α also changes the position of the maximum value of w (despite location parameter θ being kept constant), so we will write θ0 as the value of θ that provides maximum fitness for a trait value of zero.

# Library
pacman::p_load(latex2exp, ggplot2, dplyr, tidyr, viridis)

# delta; Optimal environment (the optimal temperature where fitness is maximised)
# sigma; Niche breadth (this is the variation in the fitness surface)
# alpha; Skewness of the fitness function

# Environmental Temperature
  set.seed(23)
     z = seq(0,45,0.1) # Temperature gradient
     n = 40            # Number of individuals
ind_sd = 0.5           # Standard deviation among individuals

# Simulation 1. Assume optimum is rather constant among individuals 
#parameters
delta <-  35  
sigma <-  9  + rnorm(n, 0, ind_sd) 
alpha <- -15 + rnorm(n, 0, ind_sd)

w<- list()
for(i in 1:length(alpha)){
w[[i]] <- 2*exp(-(z-delta)^2/(2*sigma[i]^2))*pnorm(alpha[i]*(z-delta)/sigma)
}

w2 <- data.frame(do.call(cbind, w))  %>% pivot_longer(everything(), names_to = "run", values_to = "fitness")  %>% mutate(z = rep(z,each = n))

my_theme <- function(){
  theme(legend.position = "none", 
           axis.title.y = element_text(size = 20), 
           axis.title.x = element_text(size = 20),
           axis.text.x = element_text(size = 18),
            axis.text.y = element_text(size = 18))
}

ggplot(w2, aes(x = z, y = fitness, color = run)) + geom_line() + theme_classic() + labs(y = "Physiological Rate", x = TeX("Temperature ($^\\circ$C)"), color = "Individual") + scale_color_viridis_d() + my_theme()+ geom_vline(xintercept = 28, linetype = "dashed", color = "red") +     geom_vline(xintercept = 18, linetype = "dashed", color = "blue")    

```
