---
title: "Thermal tolerance landscapes for populations and communities: applications in a changing world"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: false
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

# Introduction

Climate change is expected to result in warmer and more variable thermal environments globally resulting in more frequent and severe heat wave events [@barriopedro2023heat; @aglas2024recent]. Greater thermal variability will result in organisms experiencing more extreme heat wave events that exceed species tolerance limits. Exceeding physiological limits will have profound impacts on organismal growth, survival and reproduction with cascading effects on population dynamics, community composition and ecosystem services that human populations rely upon. Climate variability has already resulted in dramatic declines in pollinator abundance [@kazenel2024heat] and in crop yields globally [@zhao2017temperature; @tito2018global], but the causes underlying such declines are not fully understood. Declines are likely driven by a combination of direct effects of heat stress on plant and insect physiology, indirect effects on interactions among species -- including the microbial communities with which they interact [@trivedi2022plant] -- and how temperature interacts with other environmental stressors. Understanding the dynamic interplay between direct and indirect interactions has been hampered by the complexity of community interactions, lack of methods for measuring and predicting changes in interaction strength in diverse communities and how such interactions impact community assembly [@trivedi2022plant]. Understanding how thermal tolerance varies across populations and communities is crucial for predicting how climate change will affect ecosystems and the services they provide.

- thermal load sensitivity, damage and repair as key concepts to introduce

- What we hope to achieve in the paper.

# Species intrinsically vary in their ability to cope with heat stress

# Applying theory on thermal load sensitivity to plant and animal communities

# Microbial communities impact thermal tolerance in plants and animals

- focus on direct interactions
Interactions with symbiotic microbiota can also directly impact thermal tolerance of a focal species. For example, in plants, Márquez *et. al* @marquez2007virus describe a three-way mutualistic association between a tropical panic grass (*Dichanthelium lanuginosum*), the fungus, *Curvularia protuberata*, and a *Curvularia* virus whereby thermal tolerance of the fungus and grass is highly dependent on fungal-virus associations. Endophytic fungus cured of the virus were unable to survive extreme heating events (temperatures of 65$^\circ$C for 2 weeks). However, when the fungus were inoculated with the virus, heat tolerance was restored. Fascinatingly, the virus could be manipulated to confer heat tolerance to tomato (*Solanum lycopersicon*) demonstrating the potential for conferring thermal tolerance to other plant species. The mechanisms by which fungal-virus associations confer thermal tolerance are not yet understood but are hypothesized to involve scavenging of reactive oxygen species (ROS) by fungal endophytes generated by plant defense mechanisms in response to heat stress. 

# Heat stress can shape and reshape species interactions in communities: a view from thermal tolerance landscapes

- direct and indirect effects of heat stress on interactions
Competition among species within a community can influence a focal species' tolerance to heat stress. Competition can influence microhabitat selection and the availability of thermal refugia which can protect against thermal extremes, the diet of a focal species that can indirectly influence physiological tolerance of temperature stress, and the availability of resources that can help buffer against heat stress. For example, Li *et al.* @li2023interspecific measured the thermal tolerance landscape for three aphid species that are pests for cereal crops globally to understand whether they could predict changes in community abundance. Using a thermal death time (TDT) approach to estimate the cumulative impacts of heat stress on survival they found that one species, *Rhopalosiphum padi*, showed high heat tolerance and accumulated less injury over natural temperature fluctuations. The ability to maintain high relative abundance in the face of extreme heat may allow this species to outcompete other aphids explaining why it dominates in southern parts of its range.


Li *et al.* @li2023interspecific also found that different life-stages were affected differently by heat stress, with adults being more thermally tolerant than early developmental stages emphasising the need for models to capture such variation. However, such findings do not incorporate acclimation to heat stress or different microhabitat choice by the different species, which are both likely to be important for predicting the impacts of heat stress on community structure. It also ignores how interspecific interactions between competing aphid species will also likely mediate species abundance. It would be worth while integrating thermal tolerance landscapes with ecological models that capture interspecific interactions to better predict how heat stress will affect community structure.

# Modelling thermal tolerance landscapes to capture interactions among species

-chat with Margie and Mike about this some more. But see @zou2024time. The paper provides a framework for modelling soil plant interactions

# Methods for manipulating thermal tolerance to maintain functional interactions

*Genetic engineering*: The use of genetic engineering to confer thermal tolerance to plants and animals is a promising avenue for enhancing thermal tolerance. For example, the over-expression of heat shock proteins (HSPs) in plants has been shown to enhance thermal tolerance. HSPs are a family of proteins that are up-regulated in response to heat stress and are involved in protein folding and degradation, and the prevention of protein aggregation. CRISPR-Cas9 gene editing technology has been used to enhance thermal tolerance in plants by modifying genes involved in heat stress response.

*Selective breeding*: Artificial selection will be a useful avenue for both plants and animals to enhance thermal tolerance. 

# Applications of thermal tolerance landscapes in a changing world

## Network theory can integrate thermal tolerance landscapes with ecological models
- seems like most people are analysing communities with network theory. Some important papers @delmas2019analysing @godoy2019coexistence @poisot2021global @hui2019invade @gomez2023role @lytle2023matrix @petsopoulos2021using Should also look at this book @dale2021quantitative

have a close look at @vitali2022disruption and @riva2023toward

limitations @terry2020finding

Also this is a cool paper @lu2023food on carnivores diets by DNA sequencing fecal data

The other way to model interactions us using foundational community ecology/ species abundance models see @carrara2015inferring. We can here model the actual abundance of a given species and it would seem more flexible to incorporate thermal tolerance infor into these models. 

# Future Directions

#### *Better understanding of the mechanisms enhancing thermal tolerance in plants and animals*: We need to understand the genomic, cellular and physiological processes that confer greater thermal tolerance and help prepare organisms for future heat stress events.

#### *Modelling interactions among plants and animals to build community thermal tolerance landscapes that predict how heat stress events will affect communities*: We need to understand how thermal tolerance of individual species will affect the thermal tolerance of communities and ecosystems.

#### *Trade-offs between thermal tolerance and other traits*: We need to understand the trade-offs between thermal tolerance and other traits such as growth rate, reproduction, and competitive ability. It is fine to confer greater thermal tolerance resistance but if this comes at a cost to reproduction, growth or biomass accumulation, then it may not be beneficial. NEED EXAMPLES

#### *Behavioural responses to heat stress*: Animals behave. We need to model that. HOw do changes in behaviour impact interactions with other species and the environment. How might this impact plant productivity   

# References

```{r, performancecurve}

## See box 1 in Weiss-Lehman et al. 2022 for a description of the model. This could be applied to non-linear fitness effects based on optimal environment.

#parameters
bmax = 2  # Maximum pop growth rate lamda (think of this as optimal fitness)
 zi  = 5  # Optimal environment (the optimal temperature where fitness is maximised)
 sdi = 3  # Niche breadth (this is the variation in the fitness surface)
  xe = seq(-10,10,0.1) # Environmental gradient (self-explanatory)

lamda <- bmax*exp(-((zi-xe)/(2*sdi))^2)
plot(xe,lamda, type = "l", ylab = "Population growth rate (Lambda)", xlab = "Environmental gradient")

## Another way of doing this from Agelettas book pg 229
#parameters
 rm = 1.5  # Maximum pop growth rate lamda (think of this as optimal fitness)
 z  = 35  # Optimal environment (the optimal temperature where fitness is maximised)
 sd = 8  # Niche breadth (this is the variation in the fitness surface)
 tb = seq(0,80,0.1) # Environmental gradient (self-explanatory)
r <- rm - ((z-tb)^2/(2*sd^2))

plot(tb,r, type = "l", ylab = "Fitness", xlab = "Environmental gradient")


```

```{r, skewnormal}
# Skew normal fitness functions from https://academic.oup.com/evolut/article/73/2/145/6882140
#The parameter θ specifies the location, sigma the width, and alpha the skew of the fitness function. α > 0 gives a right-skewed function (positive skewness), α < 0 a left-skewed function, and α = 0 a symmetric Gaussian fitness function. Importantly, α also changes the position of the maximum value of w (despite location parameter θ being kept constant), so we will write θ0 as the value of θ that provides maximum fitness for a trait value of zero.

# Library
pacman::p_load(latex2exp, ggplot2, dplyr, tidyr, viridis)

# delta; Optimal environment (the optimal temperature where fitness is maximised)
# sigma; Niche breadth (this is the variation in the fitness surface)
# alpha; Skewness of the fitness function

# Environmental Temperature
  set.seed(23)
     z = seq(0,45,0.1) # Temperature gradient
     n = 40            # Number of individuals
ind_sd = 0.5           # Standard deviation among individuals

# Simulation 1. Assume optimum is rather constant among individuals 
#parameters
delta <-  35  
sigma <-  9  + rnorm(n, 0, ind_sd) 
alpha <- -15 + rnorm(n, 0, ind_sd)

w<- list()
for(i in 1:length(alpha)){
w[[i]] <- 2*exp(-(z-delta)^2/(2*sigma[i]^2))*pnorm(alpha[i]*(z-delta)/sigma)
}

w2 <- data.frame(do.call(cbind, w))  %>% pivot_longer(everything(), names_to = "run", values_to = "fitness")  %>% mutate(z = rep(z,each = n))

my_theme <- function(){
  theme(legend.position = "none", 
           axis.title.y = element_text(size = 20), 
           axis.title.x = element_text(size = 20),
           axis.text.x = element_text(size = 18),
            axis.text.y = element_text(size = 18))
}

ggplot(w2, aes(x = z, y = fitness, color = run)) + geom_line() + theme_classic() + labs(y = "Physiological Rate", x = TeX("Temperature ($^\\circ$C)"), color = "Individual") + scale_color_viridis_d() + my_theme()+ geom_vline(xintercept = 28, linetype = "dashed", color = "red") +     geom_vline(xintercept = 18, linetype = "dashed", color = "blue")    

```
