---
title: "A systems-modelling approach to predict biological responses to extreme heat"
date: "`r Sys.Date()`"
bibliography: ../bib/refs.bib
csl: ../bib/trends-in-ecology-and-evolution.csl
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "supp.html"
    theme: simplex
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
---

## Introduction

In this supplementary material, we provide extended versions, with code, of our case studies we draw upon within our manuscript. Our case studies focus on plan-insect-microbe interactions, and how these interactions are influenced by extreme heat events. We provide the code for the models we use in our case studies, and the code for the simulations we run to predict the impact of heatwaves on these interactions.

### Load packages

```{r, libraries}
#| label: libraries
# Load required libraries
install.packages("pacman")
pacman::p_load(devtools, deSolve, ggplot2, dplyr, tidyr, purrr, patchwork)
devtools::install_github('mrke/NicheMapR')

```

## Case Study 1: Predicting the impact of heatwaves on the growth of plants, microbes and insects


First, we need to get the microclimate, or simulate our own with some heatwaves. This can be done as follows:

```{r, micro_sim1}

#| label: micro_sim1
#| echo: true

# Set up environmental model (e.g., microclimate)
	microclim <- microclimate(loc = c(30, -85), 
                          month = 6,  # June
                          n = 30,     # 30 days simulation
                          shade = 0.5)  # 50% shade

# Generate temperature data for DEB models
temperature <- microclim$output$temp  # Da
```

Now that we have the temperatures each species is occupying we can include this temperature time series in our DEB models. We can simulate the growth of plants, microbes and insects under these conditions. The outputs of each can be used as input for others. First, we need to create a dynamic energy budget model for each of the three interacting species.

```{r, deb_sim1_function}
#| label: deb_sim1_function
#| echo: true

# Define DEB model for insect pest
insect_deb <- function(temp, food) {
  kappa <- 0.8  # Fraction of energy allocated to maintenance and growth
  growth <- kappa * food * exp(-((temp - 25)^2) / 10)  # Simplified growth equation
  reproduction <- (1 - kappa) * food * exp(-((temp - 30)^2) / 15)
  list(growth = growth, reproduction = reproduction)
}

# Define DEB model for plant
plant_deb <- function(temp, light, nutrient) {
  photosynthesis <- light * exp(-((temp - 20)^2) / 10)  # Simplified photosynthesis
  growth <- photosynthesis * nutrient
  respiration <- 0.1 * photosynthesis
  list(growth = growth, respiration = respiration)
}

# Define DEB model for microbe
microbe_deb <- function(temp, organic_matter) {
  growth <- organic_matter * exp(-((temp - 30)^2) / 10)
  respiration <- 0.2 * growth
  list(growth = growth, respiration = respiration)
}
```

Now, we can run the simulation for each species.


```{r, deb_sim1}
#| label: deb_sim1
#| echo: true

# Simulate interactions over 30 days
days <- 1:30
results <- data.frame(day = days, 
                      insect_growth = NA, 
                      plant_growth = NA, 
                      microbe_growth = NA)

# Initialize resources
plant_nutrients <- 1
microbe_organic_matter <- 1
light <- 1  # Assume constant light for simplicity

for (day in days) {
  temp <- temperature[day]
  
  # Insect feeds on plant
  insect_result <- insect_deb(temp, food = plant_nutrients * 0.1)
  
  # Plant photosynthesis and growth
  plant_result <- plant_deb(temp, light, nutrient = plant_nutrients)
  
  # Microbe decompose organic matter
  microbe_result <- microbe_deb(temp, organic_matter = microbe_organic_matter)
  
  # Update resources
  plant_nutrients <- plant_nutrients + plant_result$growth - insect_result$growth
  microbe_organic_matter <- microbe_organic_matter + insect_result$reproduction - microbe_result$respiration
  
  # Record results
  results$insect_growth[day] <- insect_result$growth
  results$plant_growth[day] <- plant_result$growth
  results$microbe_growth[day] <- microbe_result$growth
}

```

Now that we have the results, we can plot them to see how the growth of each species changes over time.

```{r, deb_sim1_plot}
#| label: deb_sim1_plot
#| fig-cap: "Growth of plant, insect and microbe over 30 days"
#| fig-pos: H
#| echo: true

# Plot results
library(ggplot2)
ggplot(results, aes(day)) +
  geom_line(aes(y = insect_growth, color = "Insect Growth")) +
  geom_line(aes(y = plant_growth, color = "Plant Growth")) +
  geom_line(aes(y = microbe_growth, color = "Microbe Growth")) +
  labs(y = "Growth Rate", color = "Species") +
  theme_minimal()
```